---
title: "2: R and Data Visualization Boot Camp"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Justify components of effective data visualization
2. Apply R coding skills to create effective visualizations
3. Review and/or develop skills in R coding and syntax

## Opening Discussion
What makes an effective data visualization? What are the essential components/characteristics, and what are optional but nice-to-have components/characteristics? 

Good colors & colors everyone can see; accessible and attractive color palettes
Simple and clear as possible
Labeled (especially with units) (essential)
Scale of data makes sense
Easy to understand and read for all audiences
Axis labels
Units
Legible font sizes
data point sizes - large/small enough
Title (context dependent)
Appropriate graph type
Scale of axis
Intearctive (if online and not on paper)

Come up with a consensus list among your group members. 

Compare your list to other groups' lists. 

## Exploring web resources for visualization

1. Navigate to https://www.data-to-viz.com/.
2. Think of a dataset that you or someone in your field might want to visualize. What types of variables does this dataset contain? What information might you want to convey with a visualization?

<I would like to visualze water quality data over time.>

3. Scroll to the "Explore" section of the website. Follow the flow chart to find a type of graph that would suit your data. Click for more information and a link to the R Graph Gallery. If there are multiple options for graphing your data, how might you choose between them?

<I would choose a timeseries graph and would chose between the different options based on the data I'm trying to visualize.>

4. Discuss and compare your findings with a partner.
5. Scroll further down to the "Caveats" section of the website and click "See the Collection." Choose five topics to explore (some of these are incomplete so choose another if that is the case). Write some notes about what you learned below. 

<add notes here>

6. Discuss and compare your findings with a partner.

<Boxplot - might hide sample size of each group, but is a good way to look at data. Violin plot helps show us what the different distributions look like. Density plot shows the distribution of a numeric variable.>

## A note on color palettes

In general, the ggplot base color palettes are not the most effective option for data visualization. Compiled below is a list of color palettes available in R that may serve your purposes better. Note that when working with multiple colors, you must choose between sequential, diverging, and qualitative color palettes. Under what circumstances might each be useful?

<add notes here>

RColorBrewer (package)

* http://colorbrewer2.org
* https://moderndata.plot.ly/create-colorful-graphs-in-r-with-rcolorbrewer-and-plotly/

viridis and viridisLite (packages)

* https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
* https://ggplot2.tidyverse.org/reference/scale_viridis.html

colorRamp (function; comes with base R as part of the grDevices package)

* https://bookdown.org/rdpeng/exdata/plotting-and-color-in-r.html#colorramp

LaCroixColoR (package)

* https://github.com/johannesbjork/LaCroixColoR

wesanderson (package)

* https://github.com/karthik/wesanderson

pal_futurama (function; comes in the ggsci package)

* https://nanx.me/ggsci/reference/pal_futurama.html

## Session Set Up

```{r}
getwd()

#install.packages(tidyverse)
#install.packages(dataRetrieval)
#install.packages(zoo)
#install.packages(ggrepel)

library(tidyverse)
library(dataRetrieval)
library(zoo)
library(ggrepel)
```

## Data Visualization Challenge

We will be exploring discharge data for the Eno River, aided by the [dataRetrieval package](https://cran.r-project.org/web/packages/dataRetrieval/vignettes/dataRetrieval.html). A basic ggplot is provided for you, and an image of an edited graph is provided along with hints of how to produce it. 

Work with your group to reproduce the graph. This may require dividing up tasks and sharing code as you complete each piece of editing.

```{r}
# Import data
EnoDischarge <- readNWISdv(siteNumbers = "02096500", #site number for Eno River
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-08-01",
                     endDate = "2019-07-31")

# Look at the data frame in your Environment tab.

# Renaming columns (one method of multiple)
names(EnoDischarge)[4:5] <- c("Discharge", "Approval.Code")

# dataRetrieval also includes attribute information
attr(EnoDischarge, "variableInfo") #attr is the attribute call sign; 
  # note: imperial, not metric
attr(EnoDischarge, "siteInfo")
  
# Build a ggplot
EnoPlot <- 
  ggplot(EnoDischarge, aes(x = Date, y = Discharge)) +
         geom_line() +
         xlab("Year")
print(EnoPlot)
```

Challenge: 

1. Build a plot called EnoPlot2.

2. Add a column to your data frame for discharge in meters cubed per second

hint: package dplyr in tidyverse includes a `mutate` function

3. Add a column in your data frame for a 30-day rolling mean of the metric discharge

hint: package dplyr in tidyverse includes a `mutate` function
hint: package zoo includes a `rollmean` function

4. Update your ggplot theme

hint: https://ggplot2.tidyverse.org/reference/ggtheme.html

5. Edit axes

6. Create two geom_line aesthetics, one for metric discharge and one for rolling mean of metric discharge. Color these differently.

7. Label the day that Hurricane Florence hit North Carolina. 

hint: look up this date
hint: package ggrepel includes a `geom_label_repel` function

8. Add a legend. 

hint: Google "add legend two geom layers ggplot"

```{r}  

#1. adding column 
EnoDischarge <- EnoDischarge %>%
  mutate(Discharge2 = Discharge / 35.3147)

#3. 
EnoDischarge <- EnoDischarge %>%
  mutate()


```

## Closing Discussion
In what ways was the second plot a more effective visualization than the first? 

What portions of the coding were challenging for you?
