---
title: "3: Physical Properties of Lakes"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
a
## Lesson Objectives
1. Investigate the concepts of lake stratification and mixing by analyzing monitoring data
2. Apply data analytics skills to applied questions about physical properties of lakes
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
What is a lake? How does this differ from a stream, river, or wetland?

A lake is a waterbody that has stagnant water, that can store large amounts of water. Velocity of flow in streams/rivers may be greater than velocity in lakes. Lakes could be deeper than streams or rivers. Wetlands have more vegetation than a lake.

Are streams smaller than rivers?

What are the physical properties of lakes?
- deeper than streams and rivers
- stagnant water (lentic systems)
- contained waterbody (not connected to an ocean), usually in a depression
- light that can come into the lake

## Session Set Up
```{r, message = FALSE}
# Check working directory (should be project file location)
getwd()

# load packages
library(tidyverse)
library(lubridate)

# Load data
NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

# set ggplot theme
theme_set(theme_classic())
```

## Data Wrangling and Exploration

### Investigate structure of dataset

Common steps/checks for data wrangling: 

* Date formatting
* Addressing NAs
* Subsetting and filtering 

```{r}
# Is the date column perceived as a date by R?
class(NTLdata$sampledate)
NTLdata$sampledate <- as.Date(NTLdata$sampledate, "%m/%d/%y")
class(NTLdata$sampledate)
# What does this column look like now?
#This column has the dates in yyyy/mm/dd format and uses dashes instead of slashes.

# Remove rows that contain NAs in temperature column
dim(NTLdata)
NTLdata <- NTLdata %>%
  drop_na(temperature_C)
dim(NTLdata)  

# How many observations are there for each lake?
summary(NTLdata$lakename)

# Let's choose the three lakes with data
NTLdata <- NTLdata %>%
  filter(lakename %in% c("Paul Lake", "Peter Lake", "Tuesday Lake")) # %in% means include when using pipe functions & filter

# What is another way to use the filter command to get to the same result?
NTLdata <- NTLdata %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake" | 
           lakename == "Tuesday Lake")

# Make three data frames, one for each lake
Pauldata <- filter(NTLdata, lakename == "Paul Lake")
Peterdata <- filter(NTLdata, lakename == "Peter Lake")
Tuesdaydata <- filter(NTLdata, lakename == "Tuesday Lake")

# How long did the monitoring last?
min(Pauldata$sampledate)
max(Pauldata$sampledate)
min(Peterdata$sampledate)
max(Peterdata$sampledate)
min(Tuesdaydata$sampledate)
max(Tuesdaydata$sampledate)

# Which depths are sampled in each lake?
unique(Pauldata$depth)
unique(Peterdata$depth)
unique(Tuesdaydata$depth)
# Why didn't we use the "summary" function here? Because summary will give you the mean, median, and mode, and will not give you the different depth values. Unique will give you the values of depth in each lake.

```

### Exploratory data visualization

Let's make a plot of temperatures by depth. There are a lot of points, so adding a 50 % transparency to the points helps us see where points are densely clustered together.
```{r}
TempvsDepth <- 
  ggplot(Pauldata, aes(x = depth, y = temperature_C)) + 
  geom_point(alpha = 0.5) +
  labs(y = expression("Temperature "(degree*C)), x = "Depth (m)")
print(TempvsDepth)
```

How do temperatures at the surface compare to temperatures at the mid-depths and at the bottom?
<Greater variability at the surface of the lake in terms of temperature. As lake depth increases, the temperature decreases and there's less variability in temperature.>

Let's make a few data frames that include measurements from specific depths. We will choose dates only from the 2000s for ease of interpretation.

```{r}
Pauldata2000s <- filter(Pauldata, year4 > 1999)
Pauldata.surface <- filter(Pauldata2000s, depth == 0) 
Pauldata.2m <- filter(Pauldata2000s, depth == 2)
Pauldata.3m <- filter(Pauldata2000s, depth == 3)
Pauldata.4m <- filter(Pauldata2000s, depth == 4)
Pauldata.5m <- filter(Pauldata2000s, depth == 5)
Pauldata.6m <- filter(Pauldata2000s, depth == 6)
Pauldata.7m <- filter(Pauldata2000s, depth == 7)
Pauldata.9m <- filter(Pauldata2000s, depth == 9)
Pauldata.bottom <- filter(Pauldata2000s, depth == 12)
```

Now let's plot a few of the data frames on the same ggplot. How do temperatures at the surface compare to those at 2 m depth? How do temperatures at the bottom compare to those at 9 m depth?
<This tells us that our bottom layer is mixed (hypolimnion). Warmer temperatures at surface because of interaction with the atmosphere, and also warmer temperatures at 2 meters.>

#remember to stop coding at 80 and click 'enter'

Here we are highlighting depths considered the **epilimnion** and **hypolimnion**.
```{r}
Tempplot.mixed <- 
  ggplot(Pauldata.surface, aes(x = sampledate, y = temperature_C)) +
  geom_point(color = "#6baed6", size = 1) + 
  geom_point(data = Pauldata.2m, aes(x = sampledate, y = temperature_C), 
             color = "#2171b5", size = 1) +
  geom_point(data = Pauldata.9m, aes(x = sampledate, y = temperature_C), 
             color = "#08519c", size = 1) +
  geom_point(data = Pauldata.bottom, aes(x = sampledate, y = temperature_C), 
             color = "#08306b", size = 1) +
  labs(y = expression("Temperature "(degree*C)), x = "Year")
print(Tempplot.mixed)
```

Now let's take out the 2 m and 9 m points and add in the 4 m and 6 m points. How do these compare to the surface and bottom temperatures?
<These depths are associated with the metalimnion or the thermocline. Differences in density are going to make mixing not happen.>

Here are are highlighting depths considered the **metalimnion** or **thermocline**.
```{r}
Tempplot.thermocline <- 
  ggplot(Pauldata.surface, aes(x = sampledate, y = temperature_C)) +
  geom_point(color = "#6baed6", size = 1) + 
  geom_point(data = Pauldata.4m, aes(x = sampledate, y = temperature_C), 
             color = "#2171b5", size = 1) +
  geom_point(data = Pauldata.6m, aes(x = sampledate, y = temperature_C), 
             color = "#08519c", size = 1) +
  geom_point(data = Pauldata.bottom, aes(x = sampledate, y = temperature_C), 
             color = "#08306b", size = 1) +
  labs(y = expression("Temperature "(degree*C)), x = "Year")
print(Tempplot.thermocline)
```

Why can't we use a line graph here?
<Connects between different time; will need to insert NAs if you want to use line graphs to depict this data. Otherwise you should use point data.>
```{r}
Tempplot.line <- 
  ggplot(Pauldata.surface, aes(x = sampledate, y = temperature_C)) +
  geom_line(color = "#6baed6", size = 1) + 
  geom_line(data = Pauldata.4m, aes(x = sampledate, y = temperature_C), 
             color = "#2171b5", size = 1) +
  geom_line(data = Pauldata.6m, aes(x = sampledate, y = temperature_C), 
             color = "#08519c", size = 1) +
  geom_line(data = Pauldata.bottom, aes(x = sampledate, y = temperature_C), 
             color = "#08306b", size = 1) +
  labs(y = expression("Temperature "(degree*C)), x = "Year")
print(Tempplot.line)
```

## Data Visualization and Analysis
### Creating profile graphs

The field of **limnology**, the study of inland waters, uses a unique graph format to display relationships of variables by depth in a lake (the field of oceanography uses the same convention). Depth is placed on the y-axis in reverse order and the other variable(s) are placed on the x-axis. In this manner, the graph appears as if a cross section were taken from that point in the lake, with the surface at the top of the graph.

```{r}
Pauldata.2016 <- filter(Pauldata, year4 == 2016)
Pauldata.May2016 <- filter(Pauldata, sampledate == "2016-05-17")
Pauldata.June2016 <- filter(Pauldata, sampledate == "2016-06-21")
Pauldata.July2016 <- filter(Pauldata, sampledate == "2016-07-26")
Pauldata.Aug2016 <- filter(Pauldata, sampledate == "2016-08-16")

TempprofileMay2016 <- 
  ggplot(Pauldata.May2016, aes(x = temperature_C, y = depth)) +
  geom_line(color = "#081d58") +
  geom_vline(xintercept = 4, lty = 2) + #represents the dotted line and is at 4 degrees and that is the coldest water will ever get, otherwise you'll get ice crystals
  scale_y_reverse(breaks = c(0, 2, 4, 6, 8, 10, 12)) + #reverses the y-axis
  scale_x_continuous(position = "top", limits = c(0, 30)) +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)")
print(TempprofileMay2016)

TempprofileJune2016 <- 
  TempprofileMay2016 +
  geom_line(data = Pauldata.June2016, aes(x = temperature_C, y = depth), 
            color = "#253494")
print(TempprofileJune2016)

TempprofileJuly2016 <- 
  TempprofileJune2016 +
  geom_line(data = Pauldata.July2016, aes(x = temperature_C, y = depth), 
            color = "#225ea8")
print(TempprofileJuly2016)

TempprofileAug2016 <- 
  TempprofileJuly2016 +
  geom_line(data = Pauldata.Aug2016, aes(x = temperature_C, y = depth), 
            color = "#1d91c0")
print(TempprofileAug2016)
```
<In May 2016, once the ice melts, we might expect it to be 4 degrees. It gets warm in May 2016, and heats up the surface water taht is in contact with the sun. Most of the lake is still pretty cold. Epilimnion maximum depth is between 2m - 3m. Hypolimion (highest depth of our hypolimnion) is probably 5m and we get a consistant temperature change.>

In some places, the lines are not consistently smooth. What's going on here? How does this differ from your expectations of what the data should look like?
<Epilimnion has gotten warmer as the summer has continuted. This represents temperature profiles of lakes during the summer. Lines are not smooth because of sampling error, sometimes ecosystems don't act like you want to, sampling equipment error/currents through the water.>

If you were to explain this graph to someone who didn't know anything about lakes, how would you describe it? Write your answer below. 

> 

### Mixing and stratification

Let's visualize all of the sampled dates in 2016 at once.

```{r}

Tempprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c() +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)", 
       color = "Julian Day")
print(Tempprofiles2016)

#not always going to get a consistant profile as we would expect throughout the summer.  
```

Paul Lake, like many lakes in temperate climates, is **dimictic**. What does this mean? What evidence do you see of this phenomenon in the graph? If you had data from the rest of the year, what would you expect to see?
<Dimictic - mixes twice, once in the fall and once in the spring. Fall turnover - as the air cools, you lose heat, so then lake mixes. Spring turnover - ice on the top melts and at somepoint in the spring, you'll have a water column that is all 4 degrees celcius. Mixing in the lakes is wind driven. If it's a really windy day, the lake won't mix. A completely still lake won't mix. Need a small enough density gradient so that mixing happens.>


Let's put your prediction to the test. In 1993, the lake was sampled in November. Wrangle your data to capture this date and create a profile graph of that date.
```{r}
#filtering dataset to see just November


```

Why does mixing occur in the spring and fall? What are the mechanisms that make this possible?
<add notes here>

## Closing Discussion

What are the main concepts you learned about the physical properties of lakes today? What was the evidence for these concepts in the dataset?
