---
title: "5: Physical Properties of Rivers"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Analyze seasonal and interannual characteristics of stream discharge 
2. Compare discharge patterns in different regions of the United States
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
What are the physical properties of streams and rivers? What variables might be of interest?
- flow & velocity of water (volume/time)
- velocity (distance/time) - horizontal distance
- change/meandering stream
- sediment movement
- gradient or slope for a stream/river
- bed material --> erosion --> turbidity 
- temperature 
- mixing --> turbulent & laminar flow; less mixing by depth in rivers/streams, more mixing as water moves downstream and turbulent flow


## Session Set Up

```{r, message = FALSE}
getwd()

library(tidyverse)
library(dataRetrieval)
library(cowplot)
library(lubridate)

theme_set(theme_classic()) #theme_minimal
```

## Discharge

One of the most important physical characteristics of a stream or river is **discharge**, the volume of water moving through the stream over a given amount of time. Discharge can be measured directly by measuring the velocity of flow in several spots in a stream and multiplying the flow velocity over the cross-sectional area of the stream. However, this method is effort-intensive. An easier way to approximate discharge is by developing a **rating curve** for a stream at a given sampling point. To develop a rating curve, a series of measurements of discharge and stream stage must be made together to develop a mathematical relationship. Then, subsequent discharge measurements can be calculated from a measurement of stream stage, measured by a gage that takes readings remotely. 

```{r}
EnoDischarge.ratingcurve <- readNWISdv(siteNumbers = "02096500",
                     parameterCd = c("00060", "00065"),
                     startDate = "2009-08-01",
                     endDate = "2019-07-31")

names(EnoDischarge.ratingcurve)[4:7] <- c("Discharge", "D.Approval.Code",
                                          "Gage.Height", "GH.Approval.Code")

DvsGHplot <-
  ggplot(EnoDischarge.ratingcurve, aes(x = Discharge, y = Gage.Height)) +
  geom_point(alpha = 0.5) 
print(DvsGHplot)

#interpretation of graph: have a logarithmic relationship between discharge and gage height; positive correlation over all, but as discharge increases, gage height is gaining less and less height as discharge increases. 
#if we flipped the axes, the curve would be exponential
#discharge has been empirically calculated based off the gage height

```

### Data import
We will again be using the dataRetrieval package, which allows us to access USGS hydrologic data. We will be using the hydrologic data capabilities of this package today, but the package also comes with additional functionality to access water quality data from the Water Quality Portal. 

dataRetrieval Vignette:
https://cran.r-project.org/web/packages/dataRetrieval/vignettes/dataRetrieval.html

The function `whatNWISdata` allows us to access metadata about a site, including what variables have been monitored at that site, start and end dates, and the number of samples for a given variable.

We will be analyzing data from the following rivers/streams today: 

* Eno River in North Carolina
* Verde River in Arizona (the largest perennial stream in AZ)
* Bitterroot River in Montana
* Sauk River in Minnesota (Dr. Salk's hometown)
* Nehalem River in Oregon

```{r}
EnoSummary <- whatNWISdata(siteNumbers = "02096500")
VerdeSummary <- whatNWISdata(siteNumbers = "09504000")
BitterrootSummary <- whatNWISdata(siteNumbers = "12344000")
SaukSummary <- whatNWISdata(siteNumbers = "05270500")
NehalemSummary <- whatNWISdata(siteNumbers = "14299800")

parameterCd <- c("00060")
statCd <- c("00001")
```

What data are available from the sites? Pay particular attention to the parameter codes and statistic codes.

> Site Name, longitude, latitude, datum coordinates, huc code, begin sample date, end sample date, altitudes. These metadata are hard to read because there is a lot of information and codes that we don't know/understand. Each row shows us the diffrent type of data availalbe. 

What is the period of record for mean daily discharge at each of these sites?

> Eno River: 1928-01-01 to 2019-09-10

> Verde River: 1915-06-18 to 2019-09-10

> Bitterroot River: 1937

> Sauk River: 1909-07-01 to 2019-09-10

> Nehalem River: 2001-07-26 to 2019 -09-10

The function `readNWISdv` allows us to access daily hydrologic data, which include discharge, gage height, temperature, precipitation, and pH. The function `readNWISqw` allows us to access water quality data. 

We will access the last 10 years of data for our purposes today. Alternatively, you could code `startDate = ""` to indicate you want to access all available data for the entire period of record. Then, you could filter out the years you don't want to analyze.

```{r}
# Import data
EnoDischarge <- readNWISdv(siteNumbers = "02096500",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")

VerdeDischarge <- readNWISdv(siteNumbers = "09504000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")

BitterrootDischarge <- readNWISdv(siteNumbers = "12344000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")

SaukDischarge <- readNWISdv(siteNumbers = "05270500",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")

NehalemDischarge <- readNWISdv(siteNumbers = "14299800",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")

```

### Data Wrangling
```{r}

# Renaming columns (one method of multiple)
names(EnoDischarge)[4:5] <- c("Eno.Discharge", "Eno.Approval.Code")
names(VerdeDischarge)[4:5] <- c("Verde.Discharge", "Verde.Approval.Code")
names(BitterrootDischarge)[4:5] <- c("Bitterroot.Discharge", "Bitterroot.Approval.Code")
names(SaukDischarge)[4:5] <- c("Sauk.Discharge", "Sauk.Approval.Code")
names(NehalemDischarge)[4:5] <- c("Nehalem.Discharge", "Nehalem.Approval.Code")

# Remove agency_cd column from each data frame
EnoDischarge <- select(EnoDischarge, -agency_cd)
VerdeDischarge <- select(VerdeDischarge, -agency_cd)
BitterrootDischarge <- select(BitterrootDischarge, -agency_cd)
SaukDischarge <- select(SaukDischarge, -agency_cd)
NehalemDischarge <- select(NehalemDischarge, -agency_cd)

# Join data frames
CombinedDischarge <- full_join(EnoDischarge, VerdeDischarge, by = "Date") %>%
  full_join(., BitterrootDischarge, by = "Date") %>%
  full_join(., SaukDischarge, by = "Date") %>%
  full_join(., NehalemDischarge, by = "Date")

```

### Data Visualization
```{r}
# Build a ggplot
EnoPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Eno.Discharge)) +
  geom_line() +
  ggtitle("Eno River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())
print(EnoPlot)

VerdePlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Verde.Discharge)) +
  geom_line() +
  ggtitle("Verde River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())
print(VerdePlot)

BitterrootPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Bitterroot.Discharge)) +
  geom_line() +
  ggtitle("Bitterroot River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())
print(BitterrootPlot)

SaukPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Sauk.Discharge)) +
  geom_line() +
  ggtitle("Sauk River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())         
print(SaukPlot)

NehalemPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Nehalem.Discharge)) +
  geom_line() +
  ggtitle("Nehalem River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12))
print(NehalemPlot)

CombinedPlot <- 
  plot_grid(EnoPlot, VerdePlot, BitterrootPlot, SaukPlot, NehalemPlot, 
            ncol = 1)
print(CombinedPlot)
```

How does the magnitude of discharge compare for these rivers?

> There are large differences in discharge among the rivers. Some rivers have greater peaks than other rivers. Eno has the highest peak value, but the Verde River has the most extreme from the base to the peak - it's very low and then peaks quickly. Eno River is more consistant rain fall and the discharge isn't seasonal. 

Which of these rivers have a seasonal cycle of discharge? How might you explain the presence and timing of seasonal cycles based on climatic conditions?

> The Bitterroot river and the Sauk River seem to have seasonal patterns in terms of discharge. The presence of these seasonal cycles are based on climatic conditions because it looks like the the peak discharge amount happens during the spring months when the snow melts. The Sauk River also has some increase in flow during the spring, even if it's not as large as when the snow melts. Also the Nehalem River has seasonal cycles. Snowpack melt/spring runoff is the cause for these large discharges. 

Hypothetical question: If you wanted to create a ggplot using facets instead of `plot_grid`, how would you go about wrangling your dataset differently? 

> You can use rbind() to stack the data on top of one another by site number and then can facet by site number. You need the same column titles when using rbind. 

## Defining "typical" discharge pattern
Let's take the Bitterroot River, which displays a clear seasonal cycle. Let's define the median discharge on a given day of year and overlay this onto the range experienced. We will take a longer period of record this time.
```{r}  
BitterrootDischarge.Historical <- readNWISdv(siteNumbers = "12344000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "1969-09-01",
                     endDate = "2019-08-31")

names(BitterrootDischarge.Historical)[4:5] <- c("Discharge", "Approval.Code")

BitterrootDischarge.Historical <- mutate(BitterrootDischarge.Historical, 
                                         DOY = yday(Date)) #DOY = day of year; creates a new column for day of year

BitterrootDischarge.Pattern <- BitterrootDischarge.Historical %>%
  group_by(DOY) %>% #going group by the day of the year; so it will take Jan 1 of every year and group them together
  summarise(Median.Discharge = median(Discharge), #do summary statistics
            Max.Discharge = max(Discharge),
            Min.Discharge = min(Discharge))

BitterrootPatternPlot <- 
  ggplot(BitterrootDischarge.Pattern, aes(x = DOY)) + #don't have to include y aesthetic here because we want to use a different y each time
  geom_line(aes(y = Median.Discharge)) +
  geom_line(aes(y = Max.Discharge), color = "gray") + #might be better to choose something more statistical for this instead of max & min; so choose the 90th percentile or something since max/min are more sensitive to outliers
  geom_line(aes(y = Min.Discharge), color = "gray") +  
  labs(x = "Day of Year", y = expression("Discharge (ft"^3*"/s)")) 
print(BitterrootPatternPlot)
```

Challenge: Edit the code above so that rather than plotting the minimum and maximum discharge, you are plotting the 95 % confidence interval of the discharge on a given date. Hint: look up the formula for confidence interval.

## Closing Discussion
What are your impressions of seasonal and interannual discharge for our example rivers? How might you go about interpreting discharge data for any given river?

Seasonal discharges vary a lot by location and what weather patterns they have. 
