---
title: "6: Physical Properties of Rive"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Compute recurrence intervals for stream discharge 
2. Analyze the effects of watershed disturbance on recurrence intervals and interpret results against the concept of stationarity
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
How is climate change impacting discharge in streams and rivers? What are the new and ongoing challenges faced by these impacts in watersheds?

- droughts that take up more water and therefore less discharge
- more extreme climactic events - more extreme storms/flooding
- mountanous area with lots of snowpack, climate change will push the date for snowpack melt; more rain and less snow means less predicability for flow in a river which impacts irrigation, if there's a dam, it impacts hydropower, water supply, etc. 
- estuarian zones: major changes to freshwater input, then you have a mis-balance of freshwater and salt water and salinity

## Session Set Up

```{r, message = FALSE}
getwd()

library(tidyverse)
library(dataRetrieval)
library(lubridate)


theme_set(theme_classic())
```

## Recurrence Intervals and Exceededence Probability

A **recurrence interval** is the past recurrence of an event, in this case a peak annual discharge measurement of a given magnitude. The value of a recurrence interval corresponds to the average number of years between discharge of a given magnitude. Typically the minimum amount of years required to construct a recurrence interval is 10, but 30 is more robust. Using 30 years of record gives you more information, more statistical power, the more years we have, the more "weird" years we have to consider. A recurrence interval, T, is calculated as: 

$T = (n+1)/m$

where n is the number of years and m is the ranking of an event within the observed period. We add one to n because we are computing the recurrence interval for a discharge event of a given magnitude *or greater*. 

Similarly, we can calculate an **exceedence probability**, or the probability of encountering a discharge event of a given magnitude or greater in any given year: 

$P = 1/T$

This is where the terms "100-year flood" and similar are derived. Remember this is a probability based on past occurrence, not an accurate forecast of how often we will see that event happening. When current patterns of discharge differ from past patterns, we observe **nonstationary** behavior. Nonstationarity results in events that occur more or less frequency than predicted based on the exceedence probability.

### Has Eno River dicharge displayed stationary behavior over the period of record?

Let's import discharge data for the Eno River near Durham for all available dates. Stationarirty - the river/stream behaves in the same way as it has in the past.

```{r}
EnoDischarge <- readNWISdv(siteNumbers = "02085070",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "",
                     endDate = "")

names(EnoDischarge)[4:5] <- c("Discharge", "Approval.Code")

attr(EnoDischarge, "variableInfo")
attr(EnoDischarge, "siteInfo")
  
# Build a ggplot
EnoPlot <- 
  ggplot(EnoDischarge, aes(x = Date, y = Discharge)) +
         geom_line() +
         xlab("Year")
print(EnoPlot) #data looks really variable; see pretty erratic discharge patterns in NC
```

We can then compute recurrence intervals based on the first 30 years of data. USGS uses 30 years to compute recurrence intervals.

```{r}  
EnoDischarge <- 
  EnoDischarge %>%
  mutate(Year = year(Date))

EnoRecurrence <- 
  EnoDischarge %>%
  filter(Year < 1993) %>%
  group_by(Year) %>%
  summarise(PeakDischarge = max(Discharge)) %>% #create new column and compute peak discharge
  mutate(Rank = rank(-PeakDischarge), #rank peak discharge measurements from 1-30; we want 1 to be highest peak discharge in any year the and 30 to be lowest peak discharge in any year
         RecurrenceInterval = (length(Year) + 1)/Rank, #take equation from above and compute recurrence interval; we would expect to wait 31 years to see 6210 cfs peak discharge; for other years, we may see more peak discharge amounts; in a given year, we have a 96% probability that we will see 818 peak discharge or higher in a given year
         Probability = 1/RecurrenceInterval) #compute probability which is 1/recurrence interval

EnoPeakPlot <- 
  ggplot(EnoRecurrence, aes(x = Year, y = PeakDischarge)) +
         geom_bar(stat = "identity") +
         xlab("Year")
print(EnoPeakPlot)
```

Let's display and model the relationship between peak annual disharge and recurrence interval. We can use the statistical model to compute discharge for recurrence intervals that occur above the 30-year mark.
```{r}
EnoRecurrencePlot <- 
  ggplot(EnoRecurrence, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() #+
  #scale_x_log10() 
print(EnoRecurrencePlot) #looks like a logrithmic curve; each increase in recurrence interval, increases peak discharge

Eno.RImodel <- lm(data = EnoRecurrence, PeakDischarge ~ log(RecurrenceInterval)) #linear model with log transformed recurrence interval
summary(Eno.RImodel) #very small p-value; very large R-square value, so explaining about 92% of the data with our model; any given recurrence interval can be determined based on this mathmatical relationship

#What is the discharge for a 100-year flood in this system? a 500-year flood?
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log(100) #discharge would be over 9200 cfs; dsicharge = 1332cfs (b) + 1725 cfs (m) * log(Recurrance interval) (x)
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log(500) #discharge = 12057 cfs

#these predictions assume stationarity - that the conditions will stay as they have been
```

What if we were to build a recurrence interval model for the entire period of record? How would this compare to the 30-year construction?
```{r}
EnoRecurrence.Full <- 
  EnoDischarge %>%
  group_by(Year) %>%
  summarise(PeakDischarge = max(Discharge)) %>% 
  mutate(Rank = rank(-PeakDischarge), 
         RecurrenceInterval = (length(Year) + 1)/Rank, 
         Probability = 1/RecurrenceInterval)

EnoRecurrencePlot.Full <- 
  ggplot(EnoRecurrence, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() +
  geom_point(data = EnoRecurrence.Full, color = "#02818a",
             aes(x = RecurrenceInterval, y = PeakDischarge))
print(EnoRecurrencePlot.Full) #have higher recurrence interval points; greater slope at first too; going to under predict the recurrence intervals; this shows that having more years in our data is useful

Eno.RImodel.Full <- lm(data = EnoRecurrence.Full, PeakDischarge ~ log(RecurrenceInterval))
summary(Eno.RImodel.Full)
Eno.RImodel.Full$coefficients
Eno.RImodel$coefficients

Eno.RImodel.Full$coefficients[1] + Eno.RImodel.Full$coefficients[2]*log(100)
Eno.RImodel.Full$coefficients[1] + Eno.RImodel.Full$coefficients[2]*log(500)

Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log(100)
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log(500)

```

What differences did you see for the recurrence intervals built under different periods of record? How would your prediction of flood events differ if you were to use these models for forecasting purposes? 

> For the full set of data, the recurrence intervals for 100 year flood and 500 year flood are greater than when using just the 30 year set of data. This would impact the prediction of floods because if you use the 30 year model, you will under predict the recurrence intervals of flooding. 

What would you recommend for a watershed manager seeking to build the most accurate recurrence interval model for the Eno River?

> Using the longest record you have for the stream and looking at discharges for rivers in the same area and looking at if there's a trend over time. Do we need to over estimate as we go into future, or is there a downard trend? Could use a moving window; should we maximize the data we use, or do we choose different parts? 


### Examining the effects of urbanization on discharge

Salado Creek is located in San Antonio, Texas, an area that has been rapidly urbanizing over the course of the last several decades (http://worldpopulationreview.com/us-cities/san-antonio-population/#byPopulation). Using the code above, compute recurrence intervals for Salado Creek for the first 30 years of record and for the most recent 30 years of record. How do the graphs and models differ? How do your predictions of 100- and 500-year floods differ?

```{r}
# Import data
SaladoDischarge <- readNWISdv(siteNumbers = "08178700",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "",
                     endDate = "")
names(SaladoDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(SaladoDischarge, "siteInfo")
  
SaladoPlot <- 
  ggplot(SaladoDischarge, aes(x = Date, y = Discharge)) +
         geom_line() +
         xlab("Year")
print(SaladoPlot)

# add more code here: 

#creating a year column
SaladoDischarge <- 
  SaladoDischarge %>%
  mutate(Year = year(Date))

#creating recurrence table for first 30 years 
SaladoRecurrencefirst30 <-
  SaladoDischarge %>%
  filter(Year < 1990) %>%
  group_by(Year) %>%
  summarise(PeakDischarge = max(Discharge)) %>%
  mutate(Rank = rank(-PeakDischarge),
    RecurrenceInterval = (length(Year) + 1)/Rank, 
    Probability = 1/RecurrenceInterval)

#creating recurrence table for last 30 years
SaladoRecurrencelast30 <- 
  SaladoDischarge %>%
  filter(Year > 1989) %>%
  group_by(Year) %>%
  summarise(PeakDischarge = max(Discharge)) %>%
  mutate(Rank = rank(-PeakDischarge), 
         RecurrenceInterval2 = (length(Year) + 1)/Rank,
         Probability = 1/RecurrenceInterval)

#plotting the recurrence plot
SaladoRecurrencePlot <- 
  ggplot(SaladoRecurrencefirst30, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() #+
  #scale_x_log10() 
print(SaladoRecurrencePlot) #looks like a logrithmic curve; each increase in recurrence interval, increases peak discharge

Salado.RImodel.first30 <- lm(data = SaladoRecurrencefirst30, PeakDischarge ~ log(RecurrenceInterval)) #linear model with log transformed recurrence interval
summary(Salado.RImodel.first30)

Salado.RImodel.last30 <- lm(data = SaladoRecurrencelast30, PeakDischarge ~ log(RecurrenceInterval)) 
summary(Salado.RImodel.last30)

#running equation to determine 100 and 500 year recurrence intervals 
Salado.RImodel.first30$coefficients[1] + Salado.RImodel.first30$coefficients[2]*log(100)
Salado.RImodel.first30$coefficients[1] + Salado.RImodel.first30$coefficients[2]*log(500)

Salado.RImodel.last30$coefficients[1] + Salado.RImodel.last30$coefficients[2]*log(100)
Salado.RImodel.last30$coefficients[1] + Salado.RImodel.last30$coefficients[2]*log(500)

#Plotting recurrence intervals
SaladoRecurrencePlot.Full <- 
  ggplot(SaladoRecurrencefirst30, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() +
  geom_point(data = SaladoRecurrencelast30, color = "#02818a",
             aes(x = RecurrenceInterval, y = PeakDischarge), inherit.aes = FALSE) #inherit.aes = false tells ggplot to ignore same names above (i.e. the two recurrence intervals)
print(SaladoRecurrencePlot.Full)

```

### Examining the effects of dam construction on recurrence intervals

The stream gage in the Green River near Auburn, Washington, is located directly downstream of the Howard A. Hanson Dam. The dam was built in 1961 for flood control purposes, and the reservoir now provides water supply to the city of Tacoma. How have peak discharges changed since the construction of the dam?

Using the code above, construct recurrence intervals for the periods before and after the construction of the dam. How do the graphs and models differ? How do your predictions of 100- and 500-year floods differ?

```{r}
GreenDischarge <- readNWISdv(siteNumbers = "12113000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "",
                     endDate = "")
names(GreenDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(GreenDischarge, "siteInfo")
  
GreenPlot <- 
  ggplot(GreenDischarge, aes(x = Date, y = Discharge)) +
         geom_line() +
         xlab("Year")
print(GreenPlot) #have non-stationarity in this graph of discharge, but that is because of a dam above where this stream is; you can see the large increase between 1930s and 1960s

# add more code here:

```
## Closing Discussion
This week we focused on discharge as a physical property of a stream or river. How might you use your knowledge of discharge to inform other physical processes occurring in rivers?
