---
title: "7: Water Quality in Rivers"
author: "Hydrologic Data Analysis | Cathy Chamberlin"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Analyze inorganic aspects of water quality following a watershed disturbance. 
2. Compare water quality data to federal standards.
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
What is water quality? What are some of the things it encompasses? 
<<<<<<< HEAD
 - what contaminants are or are not present in the water
 - color of water
 - smell of water
 - be able to support the biota of the water body
 - drinking water standards?
 - pH, nutrient level, ions, water chemistry
 - have to have an objective for the water: do you want to drink it? do you want to swim in it? do you want it to smell nice? 
=======
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0

## Session Set Up

```{r, message = FALSE}
getwd()

<<<<<<< HEAD
packages <- c("dataRetrieval", "tidyverse", "cowplot", "lubridate") #does the same thing as putting library in front of each package
invisible(suppressPackageStartupMessages(lapply(packages, 
                                                library, 
                                                character.only = TRUE))) #invisible makes things invisible and the messages don't come up 
=======
packages <- c("dataRetrieval", "tidyverse", "cowplot", "lubridate")
invisible(suppressPackageStartupMessages(lapply(packages, 
                                                library, 
                                                character.only = TRUE))) 
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0


theme_set(theme_classic())
```

## Specific Conductance and pH

*pH* is a measure of the acidity of water. Most natural rivers have pH close to 7 (neutral), but depending on the geology of the watershed they drain, pH can be higher or lower. Most aquatic life has difficulty living in pH less than 6. In areas high in limestone, such as much of the southern Appalachian mountain range, pH tends to be between 7.5 and 8. 

*Specific conductance* is a measure of water's ability to conduct electricity. Because conductivity depends on the number of ions in solution, specific conductance is an inexpensive estimate of how many salts are in solution. Specific conductance is dependant on the _total_ ionic strength of the water, and cannot be directly converted to concentrations of any one ion. In general though, saltier water will have higher specific conductance. Units for specific conductance in freshwater are microsiemens per centimeter at 25&deg;C.

Why are these important measures of water quality?

<<<<<<< HEAD
> to determine what can live there and what can't live there; there are only certain ranges of pH that are acceptable for certain uses of water for ecosystems, species, and/or humans

What types of disturbances can impact these parameters?

> Mining can impact these parameters as well as many other activities
=======
>

What types of disturbances can impact these parameters?

>
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0

## How does mining impact specific conductance and pH?

Let's import data from two sites in West Virginia. One site, Twelvepole Creek, experienced comparatively less mining in its water shed than the other, the Kanawha river. We will read in all of the data, but will start by looking at just data from before 1977.

```{r pull_data}
<<<<<<< HEAD
dat.raw <- readNWISqw(siteNumbers = c("03201300","03206790"), #Kanawha river & Twelvepole Creek; qw for water quality data
=======
dat.raw <- readNWISqw(siteNumbers = c("03201300","03206790"), #Kanawha river & Twelvepole Creek
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0
                     parameterCd = c("00095", "00400"), # Specific Conductance & pH
                     startDate = "",
                     endDate = "")

<<<<<<< HEAD
str(dat.raw, give.attr = FALSE) #more columns here in this dataframe than the dataframe for the daily data we downloaded last class; interested in the 'result' column as that will tell you the pH and the specific conductance values that were measured
=======
str(dat.raw, give.attr = FALSE)
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0

dat <- dat.raw %>%
  select(Site = site_no, 
         Date = sample_dt,
         Parameter = parm_cd, 
         Value = result_va) %>%
  group_by(Date, Parameter, Site) %>%
<<<<<<< HEAD
  summarize(Value = mean(Value)) %>% #summarize by mean value since on some days we have more than 1 value for pH or specific conductance each day
  spread(key = Parameter, value = Value) %>% #spread function: need a key (tells you which column you want the data to end up in), and need a value
  rename(SpC = '00095', pH = '00400') %>%
  mutate(Site = case_when(Site == "03201300" ~ "Kanawha River", #case_when funtion: like an if or statement; first part is conditional statement ~ (then) put "Kanawha River"; can also create a new column with the site name (in this case we just replaced site number because we don't want it)
=======
  summarize(Value = mean(Value)) %>%
  spread(key = Parameter, value = Value) %>%
  rename(SpC = '00095', pH = '00400') %>%
  mutate(Site = case_when(Site == "03201300" ~ "Kanawha River",
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0
                          Site == "03206790" ~ "Twelvepole Creek"))

dat.early <- dat %>%
  filter(Date < "1977-01-01")

```

Now let's compare the pH and specific conductance of the two rivers.

```{r view_pH_SpC}

pH.plot <- ggplot(dat.early, aes(x = Date, y = pH, color = Site)) +
  geom_line() +
  scale_color_manual(values = c("steelblue4", "darkorange4")) +
  ggtitle("pH of Two West Virginia Rivers 1956:1976")

print(pH.plot)

pH.plot.2 <- ggplot(dat.early, aes(x = Date, y = pH, color = Site)) +
  geom_smooth() +
  geom_point() +
  scale_color_manual(values = c("steelblue4", "darkorange4")) +
    ggtitle("pH of Two West Virginia Rivers 1956:1976")


print(pH.plot.2)

SpC.plot <- ggplot(dat.early, aes(x = Date, y = SpC, color = Site)) +
  geom_line() +
  labs(y = expression("Specific Conductance ("*mu*S * cm**-1*")")) +
  scale_color_manual(values = c("steelblue4", "darkorange4")) +
    ggtitle("Specific Conductance of Two West Virginia Rivers 1956:1976")


print(SpC.plot)
```

What differences do you see?

<<<<<<< HEAD
> Kanawha River - specific conductance is very variable; more variable than the Twelvepole creek dataset. For pH Kanawha River is more variable with pH than Twelvepole Creek, but there's also a difference in sampling frequency. If you don't measure as frequently, you won't see all the variability. pH of Kanawha River is also much higher than the pH in the Twelvepole Creek. Specific conductance - is pretty variable.  

Why might mining be causing the changes you see?

> Probably because runoff from near the mine sites picks up contaminants left behind from mining that alters the pH and specific conductance of the river. Have ions and salts that enter the river, so will have a higher specific conductance. We don't know what is in the river from the specific conductance graph, but we can say that the specific conductance in Kanawha River is higher than in Twelvepole Creek. 

What is the data frequency of these water quality samples? Why don't we have daily values? How does this inform your interpretation of the data?

> Seems like data is sampled 3 times a month. We likely don't have daily values because it is resource intensive and would require someone to go out and sample everyday. pH and specific conductance changes very little between days, so it is not worth having someone monitor it on a daily basis.  
=======
>

Why might mining be causing the changes you see?

>

What is the data frequency of these water quality samples? Why don't we have daily values? How does this inform your interpretation of the data?

>
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0
  
## What can legislative protections do for water quality?

In 1977 the Surface Mining Control and Reclamation Act (SMCRA) was passed. Take a look at how water quality changed after 1977.
```{r View_Kanawha_all}
pH.plot <- ggplot(dat, aes(x = Date, y = pH, color = Site)) +
  geom_line() +
  geom_vline(xintercept = 1977, color = "grey", alpha = 0.8, lwd = 2, lty = 2) + 
  scale_color_manual(values = c("steelblue4", "darkorange4")) +
  ggtitle("pH of Two West Virginia Rivers 1956:2000")

print(pH.plot)

pH.plot.2 <- ggplot(dat, aes(x = Date, y = pH, color = Site)) +
  geom_smooth() +
  geom_point() +
  geom_vline(xintercept = 1977, color = "grey", alpha = 0.8, lwd = 2, lty = 2) + 
  scale_color_manual(values = c("steelblue4", "darkorange4")) +
    ggtitle("pH of Two West Virginia Rivers 1956:2000")


print(pH.plot.2)

SpC.plot <- ggplot(dat, aes(x = Date, y = SpC, color = Site)) +
  geom_line() +
  geom_vline(xintercept = 1977, color = "grey", alpha = 0.8, lwd = 2, lty = 2) + 
  labs(y = expression("Specific Conductance ("*mu*S * cm**-1*")")) +
  scale_color_manual(values = c("steelblue4", "darkorange4")) +
    ggtitle("Specific Conductance of Two West Virginia Rivers 1956:2000")


print(SpC.plot)
```

Discuss with your group about the differences you see.

<<<<<<< HEAD
> pH in Kanawha River increased, but then stabilized after 1977. pH for Twelvepole Creek decreased after 1977. Because Twelvepole decreased, which we wouldn't expect. Specific conductance in both rivers was less variable and lower. 

## Were remediation efforts successful? Yes, but we will do statistical test to see the differences. 
=======
>

## Were remediation efforts successful?
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0

Let's test whether or not the yearly average concentrations in the Kanawha River are significantly different before and after 1977. 

```{r t_test}
Kanawha.yearly.averages <- dat %>%
  filter(Site == "Kanawha River") %>%
  group_by(Year = year(Date)) %>%
<<<<<<< HEAD
  select(-Site, -Date) %>% #get rid of site and date columns
  summarize_all(mean) %>%
  mutate(bin = case_when(Year %in% c(1958:1977) ~ "Pre", #20 years before the reclimation act was passed
                         Year %in% c(1978:1997) ~ "Post")) #20 years after the reclimation act was passed
=======
  select(-Site, -Date) %>%
  summarize_all(mean) %>%
  mutate(bin = case_when(Year %in% c(1958:1977) ~ "Pre",
                         Year %in% c(1978:1997) ~ "Post")) 
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0

Kanawha.early <- Kanawha.yearly.averages %>% filter(Year %in% c(1958:1977))
Kanawha.late <- Kanawha.yearly.averages %>% filter(Year %in% c(1978:1997))


histograms <- ggplot(Kanawha.yearly.averages %>% filter(!is.na(bin)), 
       aes(x = SpC, fill = bin)) +
  geom_histogram(alpha = 0.8) +
  ggtitle("Annual Averages of Specific Conductance in the Kanawha River 1958:1997") +
  labs(x = expression("Specific Conductance ("*mu*S * cm**-1*")"), 
       y = "Count") +
  scale_fill_manual(values = c("plum4", "seagreen4"),
                    name="",
                    breaks=c("Pre", "Post"),
                    labels=c("20 years before SMCRA", "20 years after SMCRA"))

print(histograms)

densityplots <- ggplot(Kanawha.yearly.averages %>% filter(!is.na(bin)), 
       aes(x = SpC, fill = bin)) +
  geom_density(alpha = 0.8) +
  ggtitle("Annual Averages of Specific Conductance in the Kanawha River 1958:1997") +
  labs(x = expression("Specific Conductance ("*mu*S * cm**-1*")"), 
       y = "Probability Density") +
  scale_fill_manual(values = c("plum4", "seagreen4"),
                    name="",
                    breaks=c("Pre", "Post"),
                    labels=c("20 years before SMCRA", "20 years after SMCRA"))

<<<<<<< HEAD
print(densityplots) #density plot is similar to a histogram and shows you the probability distribution
=======
print(densityplots)
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0

x <- (Kanawha.yearly.averages %>% filter(bin == "Pre"))$SpC
y <- (Kanawha.yearly.averages %>% filter(bin == "Post"))$SpC

qqnorm(x, las = 1); qqline(x)
qqnorm(y, las = 1); qqline(y)

cat("Length x:", length(x), "\n", "Length y:", length(y))

<<<<<<< HEAD
t.test(x, y) #yes, these are statistically significant because the p-value is very low
=======
t.test(x, y)
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0
```

How would you communicate these results to a manager? 

<<<<<<< HEAD
> The t-test shows that the specific conductance is statistically significant because the t-test resulted in a low p-value. The density plot and qqplot show that specific conductance in the Kanawha River is somewhat normally distributed as well. The t-test results suggest that the data is different for both the data before the mining act and after the mining act. Thus, the mining act was somewhat successful in mitigating high pH and/or high specific conductance.

What can you say with data visualization that you can't say with just the statistical tests?

> The data visualization plots show the distribution of the data and not just a number. 
=======
>

What can you say with data visualization that you can't say with just the statistical tests?

>
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0


## How have individual ions changed over the this timeperiod?

Let's look at the Manganese data for the Kanawha river.

```{r Manganese_data}

Manganese.raw <- readNWISqw('03201300', parameterCd = '01055')

Manganese.dat <- Manganese.raw %>%
 select(Date = sample_dt, Manganese_ugL = result_va)
  
 
Manganese.plot <- ggplot(Manganese.dat, aes(x = Date, y = Manganese_ugL)) +
  geom_line() +
  geom_vline(xintercept = 1977, color = "grey", alpha = 0.8, lwd = 2, lty = 2) + 
  labs(y = expression("Manganese ("*mu*g*l^-1*")")) +
  ggtitle("Manganese Concentrations in the Kanawha River")

print(Manganese.plot)
<<<<<<< HEAD

#plotting with points instead
Manganese.plot.point <- ggplot(Manganese.dat, aes(x = Date, y = Manganese_ugL)) +
  geom_point() +
  geom_smooth() +
  geom_vline(xintercept = 1977, color = "grey", alpha = 0.8, lwd = 2, lty = 2) + 
  labs(y = expression("Manganese ("*mu*g*l^-1*")")) +
  ggtitle("Manganese Concentrations in the Kanawha River")

print(Manganese.plot.point)
=======
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0
```

Try plotting the manganese data with points instead of lines. Which do you prefer? Why?

<<<<<<< HEAD
> I prefer the points because I find it easier to understand what is going on and the gaps between the data are visiable and easier to understand than when looking at the line graph. 

EPA drinking water standards for Manganese are 0.05 *_mg/L_*. How does the Kanawha stack up?

> The Kanawha exceeds drinking water standards for Manganese sometimes since the EPA standard is 50 micrograms/L. Before the reclimation act, we can see that the concentration of Manganese increased before the reclimation act and then decreased drastically after the act was put in place. However, even with this decrease, the concentration of Manganese in the Kanawha River was still higher than the EPA drinking water standard. 
=======
>

EPA drinking water standards for Manganese are 0.05 *_mg/L_*. How does the Kanawha stack up?

>
>>>>>>> 096cc38c2f3f68492572892a483a0b1d090a9cd0

Take some time to look at other ions as well. Some suggestions are sulfate (```00945```), nitrate (```71851```), and iron (```71885```), or use the ```whatNWISdata()``` function to find others! EPA drinking water regulations can be found here: https://www.epa.gov/ground-water-and-drinking-water/national-primary-drinking-water-regulations. Pay close attention to units! Information about parameter codes can be found in the dataframe ```parameterCdFile```.

```{r other_ion_data}


```


## Closing Discussion

Based on the data you've looked at today, what are some of the water quality impacts of mining?  What other disturbances might cause similar changes?

>

# References

Water Quality in the Kanawha-New River Basin, West Virginia, Virginia, and North Carolina, 1996–98 (USGS): https://pubs.usgs.gov/circ/circ1204/pdf/section2.pdf

National Primary Drinking Water Regulations (EPA): https://www.epa.gov/ground-water-and-drinking-water/national-primary-drinking-water-regulations

