---
title: "8: Water Quality in Rivers"
author: "Hydrologic Data Analysis | Cathy Chamberlin"
date: "Fall 2019"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Analyze biological aspects of water quality.
2. Compare water quality data to hydrologic data.
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
What are indicator species? Why are they helpful in thinking about water quality?
- If a species is really pollutant tolerant, then they might indicate that water is polluted (or vice versa)
- Indicator species: can be used to say that water quality is not good 

## Session Set Up

```{r, message = FALSE}
getwd()

packages <- c("tidyverse", 
              "dataRetrieval", 
              "cowplot", 
              "xts",
              "dygraphs", 
              "lubridate")
invisible(lapply(packages, library, character.only = TRUE)) 


theme_set(theme_classic(base_size = 12))
```

## Microbiological indicators

Bacteria is present in almost every stream on earth. Most bacteria are harmless, but some are extremely dangerous to human health.Typical sources of bacterial contamination include waste water treatment plants, animal manure lagoons, and septic systems. There are multiple ways that bacteria can be measured for water quality.

1) **Total cell counts**. One way to measure the bacteria in water is to count the number of cells under a microscope in a given volumn of water (such as 1 mL). The units for this metric are cells ml^-1^.
2) **Total coliforms**. Coliform bacteria are for the most part harmless microorganisms that live in animal intestines. Total coliforms are often used as a metric of how much raw or insufficiently treated sewage is making its way into a water body. The way these are measured are by taking a known volume of water, cultivating it in the lab, and counting the number of colonies that grow. Units for this metric are colony forming units (cfu) per 100 ml.
3) **Fecal coliforms**. Fecal coliforms are coliforms that are specifically associated with warm-blooded animals, and can be grown at higher temperatures than other coliforms. Fecal coliforms are a more precise way of estimating sewage inputs, though they are only an indicator species and not necessarily harmful. The most well known species in this group is Escherichia coli (E. coli). Units are cfu per 100 ml. 
4) **Fecal streptocci**. Fecal streptocci are a different type of bacteria that are also associated with human and animal digestive tracts. Streptocci are more widespread than coliforms. Units are cfu per 100 ml.

Why might it be helpful to have multiple ways to measure bacteria?

> Having multiple ways to measure reduces error and allows you to cross validate with other methods. Frequency of these water quality samples: really will only sample where you think there might be issues. Areas downstream of wastewater treatment plants are sampled pretty frequently.

Where and when do you think the biggest risks from harmful bacteria are?

> 

Measuring microbiogial indicators is more expensive and time consuming than some other water quality metrics. Where and when would you measure these?

>

## How have improvements in sewage treatment affected water quality?

We will import data from the Merrimack River near Lowell, Massachusetts. The Merrimack valley has a long history of industrial use, and has many mid-sized cities along the river corridor. Most of these cities have old infrastructure, and their wastewater treatment plants (WWTPs) are built next to the river and discharge directly into the water. Since the passage of the Clean Water Act, plants have either been retrofitted, or new facilities have been built to try to meet the criteria for the *point source discharges*. 

We will import a number of water quality parameters. Remember to reference ```parameterCdFile``` for metadata about the parameters.

```{r pull_data}
parameters <- parameterCdFile

dat.raw <- readNWISqw(siteNumbers = "01096550", #Merrimack River upstream of Lowell 
                     parameterCd = c("95200",  #Total cell count
                                     "31501",  #Total coliforms
                                     "31625",  #Fecal coliforms (0.7 um filter) - will pass more particles through the filter than 0.45 um filter; will have more colonies when you grow the water up
                                     "31616",  #Fecal coliforms (0.45 um filter)
                                     "31673",  #Fecal streptocci KF streptocaccus
                                     "31679"), #Fecal streptocci m-Enterococcus
                     startDate = "",
                     endDate = "")

str(dat.raw, give.attr = FALSE)

microb.dat <- dat.raw %>%
  select(Date = sample_dt, 
         Parameter = parm_cd, 
         Value = result_va) %>%
  group_by(Date, Parameter) %>%
  summarize(Value = mean(Value)) %>%
  spread(key = Parameter, value = Value) %>%
  rename(Total.coli = '31501',
         F.coli.45 = '31616', 
         F.coli.7 = '31625',
         F.strep.strep = '31673',
         F.strep.ente = '31679',
         Total.cell = '95200')

begin.date <- min(microb.dat$Date)
end.date <- max(microb.dat$Date)

Total.coli.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = Total.coli)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", y = "Total coliforms (cfu / 100 ml)") +
  theme(legend.position = "top")

print(Total.coli.plot)


F.coli.45.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.coli.45)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", 
       y = expression("Fecal coliforms filtered to 0.45 " * mu *m *  " (cfu / 100 ml)")) 

print(F.coli.45.plot)

F.coli.7.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.coli.7)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", 
       y = expression("Fecal coliforms filtered to 0.7 " * mu *m *  " (cfu / 100 ml)")) 

print(F.coli.7.plot)

F.strep.strep.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.strep.strep)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", 
       y = "Fecal streptocci, KF streptococcus (cfu / 100 ml)") 

print(F.strep.strep.plot)

F.strep.ente.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = F.strep.ente)) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", 
       y = "Fecal streptocci, m-Enterococcus (cfu / 100 ml)") 

print(F.strep.ente.plot)

Total.cell.plot <- ggplot(microb.dat, aes(x = Date)) +
  geom_line(aes(y = Total.cell), show.legend = FALSE) +
  lims(x = c(begin.date, end.date)) +
  labs(x = "", y = expression("Total cell count (# ml"^-1*")")) 

print(Total.cell.plot)

Microbes.plot <- plot_grid(Total.coli.plot +
                             labs(title = "Total Coliform", y = ""),
                           F.coli.45.plot + 
                             labs(title = "Fecal Coliform (.45)", y = "") ,
                           F.coli.7.plot + 
                             labs(title = "Fecal Coliform (.7)", y = "") ,
                           F.strep.strep.plot + 
                             labs(title = "Fecal streptocci (KF streptococcus)",
                                  y = ""),
                           F.strep.ente.plot +
                             labs(title = "Fecal streptocci (m-Enterococcus)",
                                  y = ""), 
                           Total.cell.plot + 
                             labs(title = "Total cell count", y = ""),
                           ncol = 1) #if you want to change the size of titles, will need to do this in the theme; can't do it in labs; plot.title will change the size of the plot title

print(Microbes.plot)
```

How would you use all of this data to talk about changes in microbial water quality from 1970 to 2000? What are the challenges?

> All of this data represents different forms of microbial water quality data and can therefore be used to explain the water quality. 
- hard to explain because scales are different, don't have data for entire period of record, these are also all different measures for bacteria
- total coliform is our broadest category, then fecal coliform (0.7 um), then (0.45 um); would have highest colonies of bacteria when measuring total coliform, then would have lower fecal coliform (0.7 um), and then even less fecal coliform colonies (0.45)
- can put total coliform and coliform (0.7um and 0.45um) to get a better understanding of water quality; from this graph, we can see that over time, fecal coliform decreases


Do you think water quality has improved between 1970 and 2000? Why or why not?

> Yes, water quality has improved over time between 1970 and 2000 since from the graph we can see that the amount of fecal coliform has decreased over time.

The city of Lawrence, MA is downstream of Lowell and gets its drinking water from the Merrimack River. Use the federal drinking water standard for Total Coliforms (https://www.epa.gov/ground-water-and-drinking-water/national-primary-drinking-water-regulations#Microorganisms) to comment on whether or not this water necessarily needs to be treated for bacteria.

>

## Seasonal Trends in Water Quality

```{r seasons_analysis}
monthly.obs <- microb.dat %>%
  mutate(month = month(Date)) %>%
  group_by(month) %>%
  select(-Date) %>%
  summarize_all(funs(sum(!is.na(.)))) #calculating a monthly average for all of these values

monthly.obs.strep <- microb.dat %>%
  mutate(month = month(Date)) %>%
  group_by(month) %>%
  select(-Date) %>%
  summarize_all(funs(sum(!is.na(.)))) 

coli.monthly.summaries <- microb.dat %>%
  mutate(month = month(Date)) %>%
  group_by(month) %>%
  select(month, Total.coli) %>%
  summarize_all(funs(Median = median(., na.rm = T),
                 quant25 = quantile(., .25, na.rm = T),
                 quant75 = quantile(., .75, na.rm = T))) #calculating 25th and 75th quantiles

totalcell.monthly.summaries <- microb.dat %>%
  mutate(month = month(Date)) %>%
  group_by(month) %>%
  select(month, Total.cell) %>%
  summarize_all(funs(Median = median(., na.rm = T),
                 quant25 = quantile(., .25, na.rm = T),
                 quant75 = quantile(., .75, na.rm = T))) #calculating 25th and 75th quantiles

coli.seasons <- ggplot(coli.monthly.summaries, aes(x = month)) +
  geom_ribbon(aes(ymin = quant25, ymax = quant75), alpha = 0.3) + #geom_ribbon - uses a maximum and a minimum value; this comes first in the code because we want the black line to be on top of the grey ribbon
  geom_line(aes(y = Median)) +
  scale_x_continuous(name = "Month",
                     breaks = c(1,2,3,4,5,6,7,8,9,10,11,12)) +
    labs(y = "Total coliforms (cfu / 100 ml)") 

print(coli.seasons)

totalcell.seasons <- ggplot(totalcell.monthly.summaries, aes(x = month)) +
  geom_ribbon(aes(ymin = quant25, ymax = quant75), alpha = 0.3) + #geom_ribbon - uses a maximum and a minimum value; this comes first in the code because we want the black line to be on top of the grey ribbon
  geom_line(aes(y = Median)) +
  scale_x_continuous(name = "Month",
                     breaks = c(1,2,3,4,5,6,7,8,9,10,11,12)) +
    labs(y = "Total Cell (cfu / 100 ml)") 

print(strep.seasons)
```

When are total coliform concentrations highest? Why might this be?

> Concentrations are highest in the summer; water is a lot warmer in the summer in MA, so microbes can grow a lot faster in warmer months

Does this pattern also exist for the other microbial indicators?

>

## Combined Sewer Overflows (CSOs)

Many older WWTPs combine the sanitary sewer system of a city and the storm water drainage sewers (e.g. street drains) together before treatment. Treatment occurs in tanks dug in the facilities, and treatment takes usually about a day. When there is a big rainfall event, the amount of water coming through the storm drains can overwhelm the holding capacity in the tanks, and untreated sewage can get discharged into receiving waterbodies. We will next look to see if any of the high bacterial measurements correspond to storm events.

The USGS site that we got the water quality data from does not have discharge data. Other nearby sites with discharge data include a site (```01100000```) 4 miles downstream on the Merrimack River, or a site (```01099500```) 4 miles away on the Concord River. The Merrimack River at this location drains ~4,000 square miles, and the Concord River drains ~400 square miles. 

Which site's discharge data would better indicate when the city of Lowell has exprienced a big rain event?

> The smaller watershed will be a better indicator since it responds to local events more.

What would a CSO event look like in the data?

> would see a high discharge which would also correspond with a high microbial value

```{r pull_hydrology}

discharge.raw <- readNWISdv(site = "01099500", parameterCd = c("00060"), startDate = begin.date, endDate = end.date)
str(discharge.raw, give.attr = FALSE)

discharge.dat <- discharge.raw %>%
  select(Date = Date, Discharge_cfs = X_00060_00003)
  
ggplot(discharge.dat, aes(x = Date, y = Discharge_cfs)) +
  geom_line() +
  labs(y = "Discharge (cfs)", x = "Date")
```

We will use an interactive plotting function, dygraphs to look at the discharge data and microbial data overlain on each other. Let's look first at Total Coliform.

```{r make_dygraph)}
Total_Coliform <- with(microb.dat, xts(x = Total.coli, order.by = Date)) #using the xts function from the xts package; create an xts where x = total.coli, and order it by the date with the microbe data
Discharge <- with(discharge.dat, xts(x = Discharge_cfs, order.by = Date))

DyDat <- cbind(Total_Coliform, Discharge) #use cbind to combine the data frames together; creates a matrix

dygraph(DyDat) %>% 
  dySeries("Total_Coliform", axis = "y2") %>% #axis = y2 gives you an additional y-axis
  dyAxis(name = "y", label = "Discharge (cfs)") %>%
  dyAxis(name = "y2", label = "Total Coliform (cfu per 100 ml)", valueRange = c(0, 600000)) %>% #if you don't specify value range, when you zoom in, it will rescale the y-axis; might have to do something different in the homework!!
  dyRangeSelector()
```

Look through the data series and see if you can find some times when a CSO likely occured. Create a ggplot of the 3-4 weeks surrounding the event. (Hint: use the skeleton code available below )

```{r CSO_event, eval=FALSE}

begin.date <- 
end.date <- 
  
discharge.plot <- ggplot(data =  %>% 
                   filter(), 
                 aes(x = , y = )) +
            geom_line() +
            scale_x_date(limits =  , name  = )
  
microb.plot <- ggplot(data =  %>% 
                   filter(), 
                 aes(x = )) +
            geom_point(aes(y = ))+
            scale_x_date(limits =  , name = )
  
  
plot_grid( , , ncol = 1)

```

Show your plot to your group and discuss why you think it might represent a CSO.

> could see temperature signals from the last graph we plotted. 

## Closing Discussion

We have looked at 2 different measures of water quality this week. What are some of the other important parameters we could measure for water quality? Where would we be most concerned about them?

>

# References

Merrimack River - A Comprehensive Watershed Assessment Report (Massachusetts Executive Office of Environmental Affairs): https://www.mass.gov/files/2017-07/Merrimack%20River%20Watershed%20Assessment%20Report.pdf

National Primary Drinking Water Regulations (EPA): https://www.epa.gov/ground-water-and-drinking-water/national-primary-drinking-water-regulations

