---
title: "9: Water Quality in Lakes"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Navigate and explore the LAGOSNE database and R package
2. Evaluate lake water quality using the trophic state index
3. Analyze spatial and temporal patterns of water quality across the northeast U.S.

## Opening Discussion
What are the major water quality impairments experienced in lakes?  
water quality impairments
  - nutrients (eutrophic conditions), N, P, & trace nutients
  - contaminants (toxic garbage)
  - lots of ag land leads to an increase of N & P in lakes
  - DO
  - odor/color
  - invasive species

## Session Set Up

```{r}
getwd()

library(tidyverse)
library(lubridate)
#install.packages("LAGOSNE")
library(LAGOSNE)

theme_set(theme_classic())
options(scipen = 100) #this is how to fix the problems if you are getting the e notations; usually set to a high number, but can set it to anything

lagosne_get(dest_folder = LAGOSNE:::lagos_path(), overwrite = TRUE) #will install the dataset associated with the lagos dataset; only need to do this once on a computer; then it will be loaded directly to your drive. 

```

## Getting to know the LAGOSNE database
Navigate to https://lagoslakes.org/. We will explore this website to learn about the LAGOS-NE dataset, research, and data management and use initiatives undergone by the resaerch team. 

<add any pertinent notes here>


Useful introductions to the LAGOSNE R Package can be found here: 

https://github.com/cont-limno/LAGOSNE
https://cont-limno.github.io/LAGOSNE/articles/lagosne_structure.html

```{r}
# Load LAGOSNE data into R session
LAGOSdata <- lagosne_load()
names(LAGOSdata)

# If the package installation and data download has not worked, use this code: 
load(file = "./Data/Raw/LAGOSdata.rda")

# Exploring the data types that are available
LAGOSlocus <- LAGOSdata$locus #there are 141,265 lakes that are sampled, so there are 141,265 rows of lake
LAGOSstate <- LAGOSdata$state #17 states included becasue there are 17 states included in the study
LAGOSnutrient <- LAGOSdata$epi_nutr #epilimnion nutrients - nutrients that are near the surface; data more readily available from the surface of the lake instead of the other layers of the lake; there are more than one observation at each lake

# Tell R to treat lakeid as a factor, not a numeric value
LAGOSlocus$lagoslakeid <- as.factor(LAGOSlocus$lagoslakeid)
LAGOSnutrient$lagoslakeid <- as.factor(LAGOSnutrient$lagoslakeid)
```

## Wrangling data frames in LAGOSNE

LAGOSNE is stored in several pieces, comprising metadata about given lakes (one observation per lake), metadata about each state (one observation per state), and data collected from lakes (one to many observations per lake over time). To connect observations from one data frame to the next, we need to find a common variable between the data frames.

For example, let's find out how many lakes are in each state. Note that LAGOSlocus only includes the state_zoneid, whereas  LAGOSstate connects state_zoneid for each state.

Add notes about each line of code as we go along. What does each function do?
```{r}
# Join data frames
LAGOSlocations <- left_join(LAGOSlocus, LAGOSstate, by = "state_zoneid") #take dataframe we sepcified on the left, and join on the right; by is the column that they both have in common and what you want it to join by

# Order by number of lakes
LAGOSlocations <- 
  within(LAGOSlocations, 
         state <- factor(state, levels = names(sort(table(state), decreasing=TRUE))))

LakeCounts <- ggplot(LAGOSlocations, aes(x = state)) + 
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(LakeCounts)
```

## Trophic state as a metric for water quality
Robert Carlson's 1977 paper, "A trophic state index for lakes," established the first numeric categorization system for a lake's **trophic state**, the amount of biomass a given system can sustain. Trophic state is a useful water quality metric, as it can give insight into the propensity of a system to develop algal blooms, the degree of nutrient loading in the system, and a range of other potential water quality concerns experienced by other lakes with a similar trophic state (e.g., hypoxia). Knowing the trophic state will help identify HABs as well.

To calculate the **Trophic State Index**, three variables can be used. Note these should not be used to define trophic state but as indicators of the broader condition. Comparing these values in a given lake can give insight into the broader mechanisms at play.

* *chlorophyll a concentration*, a proxy for algal (phytoplankton) biomass. Pros: direct measure of primary productivity
* *Secchi disk transparency*, a measure of water clarity. Pros: simple and cheap. Cons: may yield a high TSI in highly colored lakes and in lakes where particulate matter is comprised of non-algal material; correlates really well with other metrics we are using 
* *Total phosphorus (TP)*, a nutrient essential for growth of primary producers. Assumptions: phosphorus is the limiting nutrient for phytoplankton growth (this assumption often holds only for summer months); typically limits growth in freshwater environments; TP concentrations might oscillate throughout the year; typically use TP measurements in summer months as a proxy for growth

$TSI(Chl) = 10(6 - (2.04 - 0.68 ln Chl/ln 2))$

$TSI(SD) = 10(6 - (ln SD/ln 2))$

$TSI(TP) = 10(6 - (ln(48/TP)/ln 2))$

TSI values correspond to the following trophic states: 
**0-40**: Oligotrophic
**40-50**: Mesotrophic
**50-70**: Eutrophic
**70-100**: Hypereutrophic

### Exploring the LAGOS nutrient data frame
```{r}
dim(LAGOSnutrient)
class(LAGOSnutrient$sampledate)

LAGOStrophic <- 
  left_join(LAGOSnutrient, LAGOSlocations, by = "lagoslakeid") %>%
  select(lagoslakeid, sampledate, chla, tp, secchi, 
         gnis_name, lake_area_ha, state, state_name) %>%
  mutate(sampleyear = year(sampledate), 
         samplemonth = month(sampledate), 
         season = as.factor(quarter(sampledate, fiscal_start = 12))) %>% #lubridate function has quarter which divides data up into quarters; used fiscal_start = 12 to have it start dividing stuff into quarters from month 12, which then influences what order you put the seasons in below in the levels text; quarter knows that it needs to look at the month of the date
  drop_na(chla:secchi) #getting rid of NAs in Chl_a column through the secchi column

levels(LAGOStrophic$season) <- c("Winter", "Spring", "Summer", "Fall") #naming the seasons in the levels column, instead of 1,2,3,4; need to specify in winter, spring, summer, fall because of order we put them in

```

Let's look at observations of chl, secchi depth, and TP seasonally and over the period of study.
```{r}
chlbymonth <-  
ggplot(LAGOStrophic, 
       aes(x = as.factor(samplemonth), y = chla, color = sampleyear)) +
  geom_jitter(alpha = 0.2) + #adding noise to the plot; retaining same value for y axis and spreading out so you can see the data a bit better
  labs(x = "Month", y = expression(Chlorophyll ~ a ~ (mu*g / L)), color = "Year") +
  scale_color_viridis_c(option = "magma")
print(chlbymonth)

secchibymonth <-
ggplot(LAGOStrophic, 
       aes(x = as.factor(samplemonth), y = secchi, color = sampleyear)) +
  geom_jitter(alpha = 0.2) +
  labs(x = "Month", y = "Secchi depth (m)", color = "Year") +
  scale_color_viridis_c(option = "magma")
print(secchibymonth)

tpbymonth <- 
ggplot(LAGOStrophic, 
       aes(x = as.factor(samplemonth), y = tp, color = sampleyear)) +
  geom_jitter(alpha = 0.2) +
  labs(x = "Month", y = expression(Total ~ P ~ (mu*g / L)), color = "Year") +
  scale_color_viridis_c(option = "inferno")
print(tpbymonth)
```

What do you notice about the seasonality of these variables? If we were to characterize a lake based on the value of a given variable, how might seasonality affect our interpretations?

> Chl_a - higher in the summer months; higher light and temperature affect the amount of Chl_a in water; more measurements in the summer than in the winter; looks like we did more data sampling in 1970s - 1980s than we do now; secchi by month: can see more points in the winter months as opposed to the winter months; high chl_a concentrations, decreases secchi depth; secchi depth is opposite to Chl_a in general; phosphorous - composed of phosphate (which is dissolved), have other dissolved forms of phosphorous and particulate phosphorous; measurements seem to match up with chl_a, which isn't surprising because the amount of nutrients there are, the more chl_a there will be since those organisms need phosphorous to grow

### Calculating trophic state index
Let's add a TSI value calculated from each of the three variables to the data frame. Let's also add a column that designates the lake as oligotrophic, mesotrophic, eutrophic, or hypereutrophic based on the TSI.chl value. Make notes about the code as we go along.
```{r}
LAGOStrophic <- 
  mutate(LAGOStrophic, 
         TSI.chl = round(10*(6 - (2.04 - 0.68*log(chla)/log(2)))),
         TSI.secchi = round(10*(6 - (log(secchi)/log(2)))), #round function rounds the number to a whole number instead of leaving it as a decimal; can specify what you want it to round to, but without specifying, it automatically rounds to the nearest integer
         TSI.tp = round(10*(6 - (log(48/tp)/log(2)))), 
         trophic.class = 
            ifelse(TSI.chl < 40, "Oligotrophic", 
                   ifelse(TSI.chl < 50, "Mesotrophic",
                          ifelse(TSI.chl < 70, "Eutrophic", "Hypereutrophic")))) #calculating trophic indices for chl, secchi depth & TP; take the natural log and in R when you say 'log' it is automatically taking the natural log; if you want the other log, will need to say log10
#indeally all TSI numbers should be the same regardless of the index you use, but there is just natural variability between the lakes

LAGOStrophic$trophic.class <- 
  factor(LAGOStrophic$trophic.class,
         levels = c("Oligotrophic", "Mesotrophic", "Eutrophic", "Hypereutrophic"))

# LAGOStrophic$season <- 
#   factor(LAGOStrophic$season,
#          levels = c("Spring", "Summer", "Fall", "Winter"))

#scales::show_col(colormap(colormap = colormaps$magma, nshades=16)); can generate color palletes and grab the hex codes from this
```

Now let's compare TSI values calculated from chl, secchi depth, and TP. If these were all perfectly equivalent metrics, all points should line up on the 1:1 line. 

```{r}

chlvssecchi <- ggplot(LAGOStrophic, aes(x = TSI.chl, y = TSI.secchi)) +
  geom_point(alpha = 0.1, color = "#ec6163ff") + 
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(limits = c(0, 100)) +
  geom_abline(slope = 1, intercept = 0) 
print(chlvssecchi)
  
chlvstp <- ggplot(LAGOStrophic, aes(x = TSI.chl, y = TSI.tp)) +
  geom_point(alpha = 0.1, color = "#ec6163ff") + 
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(limits = c(0, 100)) +
  geom_abline(slope = 1, intercept = 0)  
print(chlvstp)
```

What is the observed relationship, and how does it depart from the 1:1 relationship? What do values above or below the 1:1 line tell us about the conditions present in a lake?

> 1:1 line represents every calculation based on chl-a and secchi depth, they would all fall along this line; if we were predict secchi depth, then CHL-a would be the best predictor; for phosphorous we can see similar pattern; phosphorous graph can help predict the  

### Comparing TSI across states
Here are three different ways to envision TSI across the states in the LAGOS-NE database. 
```{r}
stateTSIviolin <- ggplot(LAGOStrophic, aes(x = state, y = TSI.chl)) +
  geom_violin(draw_quantiles = 0.50)
print(stateTSIviolin) #wide violin plot - represents clustering of points at that value; skinny violin plots shows the distribution as well, but it's not as wide/distributed
  
stateTSIbar <- ggplot(LAGOStrophic, aes(x = state, fill = trophic.class)) +
  geom_bar(stat = "count", position = position_dodge(preserve = "single")) +
  theme(legend.position = "top") +
  labs(x = "", y = "Number of Observations", fill = "Trophic Class") +
  scale_fill_viridis_d(option = "magma", begin = 0.2, end = 0.8, direction = -1)
print(stateTSIbar)
  
stateTSIjitter <- ggplot(LAGOStrophic, aes(x = state, y = TSI.chl, color = season)) +
  geom_rect(xmin = -1, xmax = 19, ymin = 0, ymax = 40,
            fill = "gray90", color = "gray90") +
  geom_rect(xmin = -1, xmax = 19, ymin = 40, ymax = 50,
            fill = "gray80", color = "gray80") +
  geom_rect(xmin = -1, xmax = 19, ymin = 50, ymax = 70,
            fill = "gray70", color = "gray70") +
  geom_rect(xmin = -1, xmax = 19, ymin = 70, ymax = 100,
            fill = "gray60", color = "gray60") + #rectangles behind datapoints represent the different trophic states; x = -1 to fill up on entire plot
  geom_jitter(alpha = 0.3) +
  # geom_hline(yintercept = 40, lty = 2) +
  # geom_hline(yintercept = 50, lty = 2) +
  # geom_hline(yintercept = 70, lty = 2) +
  labs(x = "", y = "TSI(chl)") +
  scale_y_continuous(limits = c(0, 100)) +
  theme(legend.position = "top") +
  scale_color_viridis_d(option = "magma")
print(stateTSIjitter)

```
What insights do we gain from the different visualizations?

> Violin: shows the distribution in density of the points; great for getting the range and seeing the distribution

> Bar: explains some of the distribution amont each 

> Jitter:

## Closing Discussion
What factors might you expect to influence TSI scores in a given area? What are some variables in the LAGOSNE database that could help you test your hypothesis?

winter we get lower values of trophic state, summer values we get higher values; some state monitor during different times of year; some of the linear points in one line may indicate that the state is monitoring and rounding the values to integers. could also be that the state does not monitor as often or the measurements only go up by by 2 units

```{r}

write.csv(LAGOStrophic, "./Data/")

```
