---
title: "13: High Frequency Data"
author: "Hydrologic Data Analysis | Cathy Chamberlin"
date: "Fall 2019"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Determine stormflow and baseflow from high frequency flow data
2. Use a hysteresis plot to understand watershed dynamics
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion

High frequency data is usually defined as frequencies significantly lower than daily (e.g. 5-minute, 15-minute, 1 hr etc). What types of hydrological and biological processes happen on this timescale that we might want to investigate?

> discharge from fast storm events; oxygen content in water (can change via biological consumption); respiration; changing temperature --> big daily change; contaminant transport; nutrients transport/amount in water


## Session Set Up

```{r, message = FALSE}
getwd()

#install.packages("EcoHydRology")

packages <- c(
  "dataRetrieval", "tidyverse", "cowplot", "EcoHydRology", "xts", "dygraphs"
  )
invisible(lapply(packages, library, character.only = TRUE)) 

theme_set(theme_classic(base_size = 12))
```

## Pull Data

```{r download data}

dischargedatTFC2018 <- readNWISuv(
  site = "02097280", #Third Fork Creek at Woodcroft Parkway Near Blands
  parameterCd = "00060",
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns() #using dataretrieval package but using readNWISuv - uv is the symbol for instantaneous values


dischargedatEC2018 <- readNWISuv(
  site = "0208675010", #Ellerbe Creek at Club Boulevard at Durham, NC
  parameterCd = "00060", 
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns() #have a time zone code when using instantaneous data; here data is all in UTC time zone; don't have to worry about day light savings time; would need to pay attention to this time zone if dealing with biological information in projects


plot_grid(
  ggplot(dischargedatTFC2018, aes(x = dateTime, y = Flow_Inst)) +
    geom_line() + 
    scale_y_log10() +
    ggtitle("Third Fork Creek") +
    labs(x = "Date", y = "Instantaneous Flow"),
  ggplot(dischargedatEC2018, aes(x = dateTime, y = Flow_Inst)) + 
    geom_line() + 
    scale_y_log10() +
    ggtitle("Ellerbe Creek") +
    labs(x = "Date", y = "Instantaneous Flow"),
  ncol = 1
) #plotted hydrographs on log scale; want to seperate base flow from quick flow; base flow = normal flow rate

#ways water flows into streams: overland flow (through pipes or "normal" overland flow); groundwater (different depths of groundwater with different rates of leaving); fall into the river channel 
          

```

The shape and size of river hydrographs following rain events can tell us how water moves through a watershed. Once rain falls, how does water get to a stream? What would the streamflow do?

> ways water flows into streams: overland flow (through pipes or "normal" overland flow); groundwater (different depths of groundwater with different rates of leaving); fall into the river channel. Rising limb of storm is when the instantaneous flow increases; base flow is the flow that follows all of the trophs vs. quick flow which is when the flow increases during storms


_"Hydrograph separation is one of the most desperate analysis techniques in use in hydrology" - Hewlett & Hibbert 1967_


```{r baseflowseparation}
TFCbaseflow <- BaseflowSeparation(
  dischargedatTFC2018$Flow_Inst, #using instantaneous flow discharge 
  filter_parameter = 0.925, #default parameter
  passes = 3 #default parameter
  ) #this creek drains from Southeern Durham into Jordan Lake; in this table 'bt' is the base flow and 'qft' is the quick flow

TFC2018 <- cbind(dischargedatTFC2018, TFCbaseflow) #binding tables to get the date information

ggplot(TFC2018, aes(x = dateTime, y = Flow_Inst)) + 
  geom_line() +
  # scale_y_log10() + #log scale freaked out because there are lots of 0s in the data; 
  geom_line(mapping = aes(x = dateTime, y = bt), color = "darkorange4") +
  geom_line(mapping = aes(x = dateTime, y = qft), color = "steelblue4")

dygraph(
  cbind(
    Flow = with(TFC2018, xts(Flow_Inst, order.by = dateTime)), 
    Baseflow = with(TFC2018, xts(bt, order.by = dateTime)), 
    Quickflow = with(TFC2018, xts(qft, order.by = dateTime))
    )
  ) %>% 
  dyRangeSelector() #green line is total flow; quick flow is light blue and is rising slightly faster than the flow; baseflow stays pretty low until a few hours after the storm started and then baseflow increases; and after the storm, a lot of the water coming out is baseflow

Export <- TFC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_), #in pozics ct format; number of seconds since some point of time; make it numeric, so it will be a number of seconds since some point in time (will get 900 seconds if its 15 minutes); had to add NA_real_ because the timestep will give you one fewwer than what you want
         baseflowexport = bt * timestep, #mlutiply by timestep to get the amount of water in cubic feet
         quickflowexport = qft * timestep) %>%
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T), #because we added the NA value in the mutate step above, we need to tell R to remove the NA value
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)

#percentage of baseflow in 2018 Third Fork Creek (TFC)
(Export$BaseflowExport_cf/Export$TotalExport_cf) * 100 #51.9% = 52%

#percentage of quickflow in 2018 at Third Fork Creek
(Export$QuickflowExport_cf/Export$TotalExport_cf) * 100 #48.1% or about 48%
  
```

What percentage of water is exported from this watershed as baseflow? What percentage as quickflow?

> Baseflow = 51.9% and quickflow is 48.1%

Repeat this analysis for Ellerbe Creek. How do the percentages compare?

> Baseflow at Ellerbe: 43.4%; qucikflow at Ellerbe is: 56.6%

```{r}

#repeating analysis for Ellerbe Creek 

Ellerbebaseflow <- BaseflowSeparation(
  dischargedatEC2018$Flow_Inst, #using instantaneous flow discharge 
  filter_parameter = 0.925, #default parameter
  passes = 3 #default parameter
  ) #this creek drains from Southeern Durham into Jordan Lake; in this table 'bt' is the base flow and 'qft' is the quick flow

EC2018 <- cbind(dischargedatEC2018, Ellerbebaseflow)

Export_Ellerbe <- EC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_), #in pozics ct format; number of seconds since some point of time; make it numeric, so it will be a number of seconds since some point in time (will get 900 seconds if its 15 minutes); had to add NA_real_ because the timestep will give you one fewwer than what you want
         baseflowexport = bt * timestep, #mlutiply by timestep to get the amount of water in cubic feet
         quickflowexport = qft * timestep) %>%
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T), #because we added the NA value in the mutate step above, we need to tell R to remove the NA value
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)

#percentage of baseflow in 2018 at Ellerbe
(Export_Ellerbe$BaseflowExport_cf/Export_Ellerbe$TotalExport_cf) * 100 #43%

#percentage of quickflow in 2018 at Ellerbe
(Export_Ellerbe$QuickflowExport_cf/Export_Ellerbe$TotalExport_cf) * 100 #56.5%

```

What does this mean physically about the watershed?

> Ellerbe Creek likely has more overland flow than Third Flork Creek. This is likekly because Ellerbe is located closer to the city than TFC and has more impervious surfaces closer to the creek. TFC is likely near or closer to trees. Both streams are urban streams. 

## Hydrologic Flashiness

A "flashy" stream increases in flow faster following a precipitation event. 

Refering to our previous work with baseflow separation, would you expect flashiness to increase or decrease with urbanization?

> Quickflow (overland flow) will get water to a stream faster; would expect flashiness to increase with urbinzation 

What do you think "urban stream syndrom" refers to?

> hydrologic flashiness is a symptom of "urban stream system". Urban streams are very flashy because they get a lot of water coming out of pipes and directly into streams. Urban streams tend to be deeper (incised) because they tend to erode their banks faster. Also have a suite of water quality parameters that go with this.

The Richards-Baker Index is an index of how flashy a stream is. It is calculated as $$RBI = (\sum_{j = 1}^n |Q_j - Q_{j-1}| / \sum_{j = 1}^n Q_j) / Catchment Area$$


Smaller catchment area will have faster immediate changes than a larger catchment area


```{r Flashiness}
#Third Fork Creek data
TFCsitedata <- readNWISsite(site = "02097280")
TFC.catchment.size <- TFCsitedata$drain_area_va #square miles

RBI_TFC2018 <- (
  sum(abs(diff(dischargedatTFC2018$Flow_Inst))) / 
    sum(dischargedatTFC2018$Flow_Inst[-1])
  ) / TFC.catchment.size #function/equation we wrote out above is here in math terms for R

#Ellerbe Creek data
ECsitedata <- readNWISsite(site = "0208675010")
EC.catchment.size <- ECsitedata$drain_area_va #square miles

RBI_EC2018 <- (
  sum(abs(diff(dischargedatEC2018$Flow_Inst))) / 
    sum(dischargedatEC2018$Flow_Inst[-1])
  ) / EC.catchment.size

```

Which stream is flashier?

> Ellerbe creek is flashier because value for flashiness index is 0.0122 whereas value for flashiness index Third Fork Creek is 0.003. Ellerbe creek has greater differences in flashiness than Third Fork Creek. 

## Chemical Flashiness

Chemical concentrations in rivers can be either highly variable with discharge, or "chemostatic", which means the concentration barely changes with discharge. 

```{r}
CCdat <- readNWISuv(
  site = "02249500", # CRANE CREEK AT MELBOURNE, FL
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2015-01-1",
  endDate = "2016-01-01"
  ) %>%
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(CCdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()


Potdat <- readNWISuv(
  site = "01646500", # POTOMAC RIVER NEAR WASH, DC LITTLE FALLS PUMP STA
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2016-01-1",
  endDate = "2017-01-01"
  ) %>% 
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(Potdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

Which of these rivers is more chemostatic?

>

## Hysteresis Loops

Hysteresis occurs when the concentration varies with discharge in a different relationship on the rising and falling limbs of a storm. This can be caused by _flushing_ behavior or _dilution_ behavior. 

```{r}
dygraph(
  cbind(
    Flow = xts(Potdat$Flow_Inst, order.by = Potdat$dateTime), 
    Nitrate = xts(Potdat$Nitrate_mgl, order.by = Potdat$dateTime)
    )
  ) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector()
```

Look at the storm that occured around February 1st. Do you expect this storm to be a flushing or a diluting storm?

>

```{r}
PotomacStorm <- Potdat %>%
  filter(dateTime > "2016-01-31" & dateTime < "2016-02-14") 

ggplot(PotomacStorm, aes(x = Flow_Inst, y = Nitrate_mgl, color = dateTime)) +
  geom_point() 


```

Did this storm exhibit clockwise or counterclockwise hysteresis?

>

Which direction do hysteresis loops go for diluting storms?

>

Which direction do hysteresis loops go for flushing storms?

>

## Closing Discussion

What hydrological processes did we investigate today? 

>

What can high-frequency data be used for that daily data may not be able to capture?

> 
