---
title: "15: Mapping"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Define the basic components of spatial mapping
2. Create maps of water quality data in R
3. Analyze and communicate the findings of spatial analysis using mapping tools

## Opening Discussion

What are some examples of spatial data in the aquatic sciences? Why might you want to map these?
 - change over time
 - ability to view
 - concentration gradients
 - catchment characteristics that might influence water quality and quantity
 - different features of a river (i.e. depth)
 - can see where contaminants are entering the water
 - mapping snow pack

## Spatial visualization

The geometry of a given spatial data point consists of coordinates in 2-, 3-, or 4-dimensional space. These dimensions are: 

* **x**: longitude (required)
* **y**: latitude (required)
* **z**: altitude (optional)
* **m**: measurement (optional) - could color point based on discharge and/or other measurements

A feature (object) most often falls into one of these three categories (more complex examples exist; see the `sf` package documentation for more): 

* **Point**: specific point will have two numbers (longitude and latitude); x, y coordinates; fit pretty easily into a tidy dataset
* **Line**: connection of points; can be connection of two points; would have more of a list and have multiple values
* **Polygon**: average area of the points; connection of lines that creates a shape; you'll see a centroid position which is a specific lat/long measurement


## Mapping in R

Conducting spatial visualization in R presents several benefits: 

1. R is an open-source software, making code and output accessible without a software license. Closed-source software such as ArcGIS offers more complex functionality, but for many purposes we can use R to the same effect. 
2. Community-sourced packages are improving the functionality of R's spatial capabilities all the time. 
3. Coding in R replaces traditional click-based programming in ArcGIS, moving toward reproducible workflow and data analysis pipelines.
4. Spatial tools in R are integrated with other statistics, visualization, and data science tools that you may already use. 

We will be using the `sf` ("simple features") package today. A helpful vignette for this package can be found here: https://r-spatial.github.io/sf/articles/sf1.html. We will also be using `maps`, which includes several useful layers of political geography around the world. We will be using their `map` function to plot outlines of U.S. states.

## Session Set Up

```{r, message = FALSE}
getwd()

library(tidyverse)
library(lubridate)
library(cowplot)
library(LAGOSNE)
library(sf)
library(maps)
library(gganimate)
#install.packages("rgeos")
#install.packages("gifski")
#install.packages("png")
#install.packages("transformr")

theme_set(theme_classic())
options(scipen = 100)

# Load LAGOSNE data into R session
LAGOSdata <- lagosne_load()
# If the lagosne_get function has not worked, use this code: 
load(file = "./Data/Raw/LAGOSdata.rda")


```
## Generating spatial data frames

We will be using the `st_as_sf` function today, which converts an object to a simple features object. Notice in the "states" object contains a column called "geometry", which contains a list of coordinates. Notice each cell in this column is a list, which can be expanded to look at the different coordinates used to draw a polygon. We plot the polygons with the `geom_sf` function. 

```{r}
# generate a map of U.S. states
states <- st_as_sf(map(database = "state", plot = TRUE, fill = TRUE, col = "white"))

# filter only states that are included in the LAGOSNE database
states.subset <- filter(states, ID %in% 
                          c("minnesota", "iowa", "wisconsin", "illinois",
                            "missouri", "michigan", "indiana", "ohio",
                            "pennsylvania", "new york", "new jersey",
                            "connecticut", "new hampshire", "rhode island",
                            "massachusetts", "vermont", "maine"))
# visualize state plot
LAGOSstateplot <- ggplot(states.subset) +
  geom_sf(fill = "white")
print(LAGOSstateplot)
```

## Data wrangling
```{r}
# load LAGOSNE data frames
LAGOSlocus <- LAGOSdata$locus
LAGOSstate <- LAGOSdata$state
LAGOSnutrient <- LAGOSdata$epi_nutr
LAGOSlimno <- LAGOSdata$lakes_limno

# Create a data frame to visualize secchi depth
LAGOScombined <- 
  left_join(LAGOSnutrient, LAGOSlocus) %>% #will chose columns that you might not want it to join by if they have the same name; be careful about what your joins are doing! should do step by step so that you can see what is happening to the data
  left_join(., LAGOSlimno) %>%
  left_join(., LAGOSstate) %>%
  filter(!is.na(state)) %>%
  select(lagoslakeid, sampledate, secchi, lake_area_ha, maxdepth, nhd_lat, nhd_long, state)
``` 

Notice that in the absence of specifying specific columns to join by, the `left_join` function will choose columns itself. The resulting data frame has the same amount of rows as the LAGOSnutrient data frame, minus any observations that listed state as NA. Be careful when relying on this functionality! Always double check your final data frame to make sure it contains the correct data and that your joins have proceeded as planned.

Let's create a new data frame of average secchi depth for each lake. Notice that in the `summarise` function (also can use `summarize`) we also compute the "mean" of max depth, latitude, and longitude for each lake. These should all be the same for every observation at a given lake, so taking the mean just uses that one value.
```{r}
secchi.summary <- LAGOScombined %>%
  group_by(lagoslakeid) %>%
  summarise(secchi.mean = mean(secchi),
            area = mean(lake_area_ha),
            depth = mean(maxdepth),
            lat = mean(nhd_lat), 
            long = mean(nhd_long)) %>%
  drop_na()
```

## A Word on Projections

The Earth is round, meaning that if we want to make a map in 2d space we need to make a projection. This becomes a particular issue when we map large areas, because the edges of the map become distorted the further away we get from the center. A great illustration of projections and their pitfalls can be found here:  https://source.opennews.org/articles/choosing-right-map-projection/. 

For today, we will use the EPSG projection 4326, also known by WGS 84. This projection is the reference system for the Global Positioning System (GPS) and functions well for the LAGOSNE database.

Projections are used to change the way the earth is viewed; distorting the view to stretch out the Earth based on the projection since the Earth is round and we are putting information on a 2-D plane.

## Plotting secchi depths across LAGOSNE lakes
```{r}
secchi.spatial <- st_as_sf(secchi.summary, coords = c("long", "lat"), crs = 4326) #projection we are using today; we are using lat and long coordinates

Secchiplot <- ggplot() +
  geom_sf(data = states.subset, fill = "white") + #states first because we are using a white fill, so if we did secchi points first, we wouldn't see them
  geom_sf(data = secchi.spatial, aes(color = secchi.mean), 
          alpha = 0.5, size = 1) +
  scale_color_viridis_c() +
  labs(color = "Average Secchi Depth (m)") +
  theme(legend.position = "top")
print(Secchiplot) #clusters of secchi depth measurements in lakes where people sample

```

## Secchi depth mapping challenge. 
Your turn! Notice that Maine has a large number of lakes with secchi depth measurements as well as a large range in secchi depth. Let's zoom in on this state for a closer look. 

1. Filter the states and secchi depth datasets so that they contain Maine only. For the secchi depth dataset, create a summary dataset with just the mean secchi depth. 

```{r}

#filtering for ME data only
Maine_secchi <- LAGOScombined %>%
  filter(state == "ME")

#Maine subset dataset
Maine_subset <- states.subset <- filter(ID == "maine")

#visualize state plot
Mainestateplot <- ggplot(Maine_subset) +
  geom_sf(fill = "white")
print(Mainestateplot)

#ME secchi summary data
ME.secchi.summary <- Maine_secchi %>%
  group_by(lagoslakeid) %>%
  summarise(secchi.mean = mean(secchi),
            area = mean(lake_area_ha),
            depth = mean(maxdepth),
            lat = mean(nhd_lat), 
            long = mean(nhd_long)) %>%
  drop_na()

#secchi spatial info for ME
secchi.spatial.ME <- st_as_sf(ME.secchi.summary, coords = c("long", "lat"), crs = 4326)

#ggplot of data
  ME.Secchiplot <- ggplot() +
  geom_sf(data = Maine_subset, fill = "white") +
  geom_sf(data = secchi.spatial.ME, aes(color = secchi.mean), 
          alpha = 0.5, size = 1) +
  scale_color_viridis_c() +
  labs(color = "Average Secchi Depth (m)") +
  theme(legend.position = "top")
print(ME.Secchiplot)


```

2. Create a plot of mean secchi depth for lakes in Maine, with mean secchi depth designated as color and the lake area as the size of the dot. Remember that you are using size in the aesthetics and should remove the size = 1 from the other part of the code. Adjust the transparency of points as needed.

```{r}
#plot of mean secchi depths for lakes in ME
ME.mean.secchi.plot <- ggplot() +
  geom_sf(data = Maine_subset, fill = "white") +
  geom_sf(data = secchi.spatial.ME, aes(color = secchi.mean, size = area),
          alpha = 0.5) +
  scale_color_viridis_c() +
  labs(color = "Average Secchi Depth (m)", size = "Area") +
  theme(legend.position = "top")
print(ME.mean.secchi.plot)

```

3. Create a second plot, but this time use maximum depth of the lake as the size of the dot. 

```{r}

```

4. Plot these maps in the same plot with the `plot_grid` function. Don't worry about adjusting the legends (if you have extra time this would be a good bonus task).

```{r}

```

5. What relationships do you see between secchi depth, lake area, and lake depth? Which of the two lake variables seems to be a stronger determinant of secchi depth? (you can make a scatterplot or run a regression to test this if time remains)

> 

## Visualizing secchi depth over time

We might want to visualize how monitoring efforts and secchi depths have changed over time. Below we will divide each secchi depth measurement by the decade in which it was taken (this can be an effective way to reduce long time series). 
```{r}
# add a "decade" column 
LAGOScombined <- LAGOScombined %>%
  mutate(decade = floor_date(sampledate, years (10)), #identifies the decade this sample is in (gives you "1980-01-01")
         decade = year(decade)) #do another mutate in order to just get the year "1980" as an output

# create a new summary data frame, with each lake divided by decade
secchi.summary.decade <- LAGOScombined %>%
  group_by(lagoslakeid, decade) %>% #so that each lake id has every decade they were sampled
  summarise(secchi.mean = mean(secchi),
            lat = mean(nhd_lat), 
            long = mean(nhd_long)) %>%
  drop_na()

# create a simple features object
secchi.decade.spatial <- st_as_sf(secchi.summary.decade, coords = c("long", "lat"), crs = 4326)

# plot the data with decades as separate facets
Secchiplot <- ggplot() +
  geom_sf(data = states.subset, fill = "white") + # put state sf first
  geom_sf(data = secchi.decade.spatial, aes(color = secchi.mean), 
          alpha = 0.5, size = 1) +
  facet_wrap(vars(decade), ncol = 5) + #creating a facet for each decade
  scale_color_viridis_c() +
  labs(color = "Average Secchi Depth (m)") +
  theme(legend.position = "top")
print(Secchiplot)
```

Faceting can be an effective way to visualize data over time. But, R has additional functionality that we can visualize these changes over time in one single graph, with the package `gganimate`. Let's build an animated plot and create a GIF. 

Note: we have installed and loaded `gganimate`, but RStudio may prompt you to install additional packages. Do this as needed until the code runs without error. 

```{r}
Secchi.Animation <- ggplot() +
  geom_sf(data = states.subset, fill = "white") +
  geom_sf(data = secchi.decade.spatial, aes(color = secchi.mean), 
          alpha = 0.5, size = 1) +
  scale_color_viridis_c() +
  theme(legend.position = "top") + #this code is just regular ggplot code, should all be recognizable
  #gganimate code here:
  labs(title = 'Decade: {closest_state}', color = "Average Secchi Depth (m)") + #title is animated, 'state' means the condition
  transition_states(decade, state_length = 1, transition_length = 0) + #the thing that will be changing will be the decade. The time it spends between states (transition_length) will be 0. Each condition will last for 1 unit of time (state_length)
  enter_appear() +
  exit_disappear() 

anim_save('Secchi.Animation.gif', animate(Secchi.Animation, height = 400, wid = 600, renderer = gifski_renderer(loop = TRUE)))

```

