---
title: "Assignment 5: Water Quality in Lakes"
author: "Caroline Watson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on water quality in lakes

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single HTML file.
8. After Knitting, submit the completed exercise (HTML file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A05_Salk.html") prior to submission.

The completed exercise is due on 2 October 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, lubridate, and LAGOSNE packages. 
3. Set your ggplot theme (can be theme_classic or something else)
4. Load the LAGOSdata database and the trophic state index csv file we created on 2019/09/27.

```{r setup}
#checking working directory
getwd()

#loading packages
suppressMessages(library(tidyverse))
library(lubridate)
library(LAGOSNE)
library(viridis)

#setting ggplot theme
theme_set(theme_classic())
options(scipen = 100)

# Load LAGOSNE data into R session
lagosne_get(dest_folder = LAGOSNE:::lagos_path(), overwrite = TRUE)

#loading LAGOS dataset
LAGOSdata <- lagosne_load()
names(LAGOSdata)

#loading trohic state index csv file
LAGOS.trophic <- read.csv("./Data/LAGOStrophic.csv")

```

## Trophic State Index

5. Similar to the trophic.class column we created in class (determined from TSI.chl values), create two additional columns in the data frame that determine trophic class from TSI.secchi and TSI.tp (call these trophic.class.secchi and trophic.class.tp). 

```{r}

#creating trophic.class.secchi column in datafram
LAGOStrophic <- 
  mutate(LAGOS.trophic, 
         TSI.chl = round(10*(6 - (2.04 - 0.68*log(chla)/log(2)))),
         TSI.secchi = round(10*(6 - (log(secchi)/log(2)))),
         TSI.tp = round(10*(6 - (log(48/tp)/log(2)))), 
         trophic.class = 
            ifelse(TSI.chl < 40, "Oligotrophic", 
                   ifelse(TSI.chl < 50, "Mesotrophic",
                          ifelse(TSI.chl < 70, "Eutrophic", "Hypereutrophic"))), 
         trophic.class.secchi = 
           ifelse(TSI.secchi < 40, "Oligotrophic", 
              ifelse(TSI.secchi < 50, "Mesotrophic", 
                     ifelse(TSI.secchi < 70, "Eutrophic", "Hypereutrophic"))), 
         trophic.class.tp = 
           ifelse(TSI.tp < 40, "Oligotrophic", 
                  ifelse(TSI.tp < 50, "Mesotrophic", 
                         ifelse(TSI.tp < 70, "Eutrophic", "Hypereutrophic"))))

```


6. How many observations fall into the four trophic state categories for the three metrics (trophic.class, trophic.class.secchi, trophic.class.tp)? Hint: `count` function.

```{r}

#number of observations in trophic.class
LAGOStrophic %>% count(trophic.class) #

#number of observations in trophic.class.secchi
LAGOStrophic %>% count(trophic.class.secchi)

#number of observations in trophic.class.tp
LAGOStrophic %>% count(trophic.class.tp)

```

7. What proportion of total observations are considered eutrohic or hypereutrophic according to the three different metrics (trophic.class, trophic.class.secchi, trophic.class.tp)?

```{r}
#proportion of tophic.class
LAGOStrophic %>% count()


```

Which of these metrics is most conservative in its designation of eutrophic conditions? Why might this be?

> 

Note: To take this further, a researcher might determine which trophic classes are susceptible to being differently categorized by the different metrics and whether certain metrics are prone to categorizing trophic class as more or less eutrophic. This would entail more complex code.

## Nutrient Concentrations

8. Create a data frame that includes the columns lagoslakeid, sampledate, tn, tp, state, and state_name. Mutate this data frame to include sampleyear and samplemonth columns as well. Call this data frame LAGOSNandP. 

```{r}
#creating LAGOS dataframes
LAGOSlocus <- LAGOSdata$locus
LAGOSstate <- LAGOSdata$state
LAGOSnutrient <- LAGOSdata$epi_nutr

#joining state and nutrient dataframes
LAGOSlocations <- left_join(LAGOSlocus, LAGOSstate, by = "state_zoneid")

#joining LAGOS locations and nutrient dataframs
LAGOSNandP <- left_join(LAGOSlocations, LAGOSnutrient, by = "lagoslakeid") %>%
  select(lagoslakeid, sampledate, tn, tp, state, state_name) %>%
  mutate(sampleyear = year(sampledate), 
         samplemonth = month(sampledate))

```

9. Create two violin plots comparing TN and TP concentrations across states. Include a 50th percentile line inside the violins.
```{r}
#violin plot comparing TN across states
TNstateviolin <- ggplot(LAGOSNandP, aes(x = state, y = tn)) +
  geom_violin(draw_quantiles = 0.50) +
  labs(x = "State", y = "Total Nitrogen Concentration (mg/L)")
print(TNstateviolin)

#violin plot comparing TP across states
TPstateviolin <- ggplot(LAGOSNandP, aes(x = state, y = tp)) +
  geom_violin(draw_quantiles = 0.50) +
  labs(x = "State", y = "Total Phosphorous Concentration (mg/L)")
print(TPstateviolin)
```

Which states have the highest and lowest median concentrations?

> TN: Iowa has the highest median concentration and Vermont has the lowest median concentration.

> TP: Illinois has the highest median concentration and Maine has the lowest median concentration.

Which states have the highest and lowest concentration ranges?

> TN: Iowa has the highest concentration range and New Hampshire and Vermont have the lowest concentration ranges.

> TP: Illinois and Minnesota have the highest concentration ranges and Connecticut and Pennslyvania have the lowest concentration ranges. 

10. Create two jitter plots comparing TN and TP concentrations across states, with samplemonth as the color. Choose a color palette other than the ggplot default.

```{r}
#jitter ggplot of TN concentrations across states
TNstate_jitter <- ggplot(LAGOSNandP, aes(x = state, y = tn, color = samplemonth)) +
 geom_jitter(alpha = 0.3) +
 labs(x = "State", y = "Total Nitrogen Concentration (mg/L)", color = "Month") +
 scale_color_viridis_c(option = "magma", begin = 0.2, end = 0.8) +
 theme(legend.position = "top")
print(TNstate_jitter)

#jitter ggplot of TP concentrations across states
TPstate_jitter <- ggplot(LAGOSNandP, aes(x = state, y = tp, color = samplemonth)) +
 geom_jitter(alpha = 0.3) +
 labs(x = "State", y = "Total Phosphorus Concentration (mg/L)", color = "Month") +
 scale_color_viridis_c(option = "magma", begin = 0.2, end = 0.8) +
 theme(legend.position = "top")
print(TPstate_jitter)

```

Which states have the most samples? How might this have impacted total ranges from #9?

> TN: Iowa has the most samples, which is likely why Iowa had the highest median concentration and highest range. 

> TP: Illinois and Minnesota look like they have the most samples. This definitely impacted the highest median concentration (Illinois) and highest concentration ranges (Illinois and Minnesota). 

Which months are sampled most extensively? Does this differ among states?

> TN: The summer to fall months are sampled most extensively. Yes, this does differ among states as some states have samples in the winter (i.e. Wisconsin).

> TP: The summer and fall monts look as if they are sampled most extensively. However, some states have samples from the winter months and fall months. 

11. Create two jitter plots comparing TN and TP concentrations across states, with sampleyear as the color. Choose a color palette other than the ggplot default.

```{r}
#jitter plot of TN across state with sampleyear as color
TNstate_jitter2 <- ggplot(LAGOSNandP, aes(x = state, y = tn, color = sampleyear)) +
 geom_jitter(alpha = 0.3) +
 labs(x = "State", y = "Total Nitrogen Concentration (mg/L)", color = "Year") +
 scale_color_viridis_c(option = "magma", begin = 0.2, end = 0.8) +
 theme(legend.position = "top")
print(TNstate_jitter2)

#jitter plot of TP across state with sampleyear as teh color 
TPstate_jitter2 <- ggplot(LAGOSNandP, aes(x = state, y = tp, color = sampleyear)) +
 geom_jitter(alpha = 0.3) +
 labs(x = "State", y = "Total Phosphorus Concentration (mg/L)", color = "Year") +
 scale_color_viridis_c(option = "magma", begin = 0.2, end = 0.8) +
 theme(legend.position = "top")
print(TPstate_jitter2)

```


Which years are sampled most extensively? Does this differ among states?

> TN: Years 1980s - 2016 were sampled most frequently. This does differ a little among states, with CT having some sampling occuring between 1950s - 1975. 

> TP: The years 1975 - 2016 were sampled most frequently for total phosphorus. However, this does differ among states as some states, such as MN, CT, ME, and VT have samples from the 1950s (or earlier) included in the dataset. 

## Reflection
12. What are 2-3 conclusions or summary points about lake water quality you learned through your analysis?

>

13. What data, visualizations, and/or models supported your conclusions from 12?

> 

14. Did hands-on data analysis impact your learning about water quality relative to a theory-based lesson? If so, how?

>

15.	How did the real-world data compare with your expectations from theory?

>
